---
title: "Cleaning and standardizing spectra"
author: "Jose Eduardo Meireles"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Cleaning and standardizing spectra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
source("_common.R")
library("spectrolab")
```

Raw leaf spectra are rarely analysis-ready. They carry sensor noise, they arrive
on whatever band grid the instrument used, and they differ in overall brightness
because of illumination geometry, leaf angle, and specular glare that have
nothing to do with the biochemistry you care about. This vignette walks through
`spectrolab`'s tools for fixing those three problems -- **smoothing**,
**resampling**, and **normalizing** (plus **derivatives** along the way) -- and
closes with the order in which to apply them.

Every transform here takes a `spectra` object and returns one, and every one
**requires strictly increasing bands**. If your data still carries overlapping
sensor ranges, splice them with `match_sensors()` first (see the
[Matching sensor overlap](match_sensors.html) vignette).

We use the simulated `spec_matrix_example` dataset throughout -- 50 spectra,
400-2500 nm, with 10 species encoded in the sample names. Because it is
simulated, it is convenient for illustration but is not a substitute for real
measurements.

```{r, message = FALSE}
s = as_spectra(spec_matrix_example, name_idx = 1)   # simulated: 50 samples, 10 species
s

one = s[1, ]   # a single spectrum, for the before/after smoothing panels
```

## Smoothing

Smoothing suppresses band-to-band noise while preserving the shape of the
spectrum. `smooth()` dispatches to four methods via its `method` argument:

| `method`          | function             | idea |
|-------------------|----------------------|------|
| `"gaussian"` (default) | `smooth_fwhm()` | resample onto the same bands using a wider Gaussian FWHM (see [Resampling](#resampling-to-a-common-grid) below) |
| `"spline"`         | `smooth_spline()`    | fit `stats::smooth.spline()` per sample |
| `"moving_average"` | `smooth_moving_avg()`| simple moving average over `n` adjacent bands |
| `"sgolay"`         | `smooth_sgolay()`    | Savitzky-Golay filter: a local polynomial fit over a moving window (requires the `signal` package) |

```{r, fig.height=3.5, fig.width=7, eval = requireNamespace("signal", quietly = TRUE)}
s_gauss = smooth(one, method = "gaussian")
s_sg    = suppressMessages(smooth(one, method = "sgolay"))

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(one, lwd = 1, col = spec_muted, main = "Gaussian (default)")
plot(s_gauss, add = TRUE, col = spec_pal[6], lwd = 1)

plot(one, lwd = 1, col = spec_muted, main = "Savitzky-Golay")
plot(s_sg, add = TRUE, col = spec_pal[1], lwd = 1)

par(oldpar)
```

### The Savitzky-Golay filter in more detail

`smooth_sgolay(x, p, n)` fits a degree-`p` polynomial over a moving window of
`n` bands and replaces the center value with the fitted one. `n` must be odd
and greater than `p`; the default, `n = p + 3 - p %% 2`, is deliberately small
(5 bands, for the default `p = 3`) so it does very little smoothing until you
widen the window:

```{r, fig.height=3.5, fig.width=7, eval = requireNamespace("signal", quietly = TRUE)}
s_narrow = smooth_sgolay(one, p = 3, n = 5)    # default-sized window
s_wide   = smooth_sgolay(one, p = 3, n = 51)   # a much wider window

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(one, lwd = 1, col = spec_muted, main = "n = 5 (default-sized)")
plot(s_narrow, add = TRUE, col = spec_pal[1], lwd = 1)

plot(one, lwd = 1, col = spec_muted, main = "n = 51")
plot(s_wide, add = TRUE, col = spec_pal[1], lwd = 1)

par(oldpar)
```

Unlike a simple moving average, the Savitzky-Golay filter has full-length output
at the edges (no bands are dropped there) because its polynomial fit is defined
at every point in the window, including the boundary -- see the
[Gotchas](#smoothing-gotchas) below for how `smooth_moving_avg()` differs.

### Smoothing gotchas

* **The `signal` package is required for `smooth_sgolay()` and
  `deriv_spectra()`**, but it is only in `Suggests` -- if it is not installed
  you get a clear error telling you to install it, rather than a cryptic one.
* **The method name is `"sgolay"`**, not `"savitzky_golay"` or `"savgol"` --
  matching the naming of the underlying `signal::sgolay()`/`sgolayfilt()`
  functions it wraps.
* **`smooth_moving_avg()` can silently drop bands.** Its moving average produces
  `NA` at the edges (there are not enough neighbors to average there), and those
  edge bands are then *removed* from the result -- your output can have fewer
  bands than the input. The dropped values are stashed into metadata
  (`save_bands_to_meta = TRUE`, the default) so they are not lost, just
  relocated. `smooth_sgolay()` does not have this issue.
* **Choose the Savitzky-Golay window (`n`) relative to your spectral
  resolution**, not in the abstract: 5 bands means something very different on
  1 nm-resolution data than on 10 nm data. The default is intentionally small;
  widen it deliberately, and check the result visually as we did above.

## Derivatives and the red edge

`deriv_spectra(x, order)` computes the order-`order` derivative of each spectrum
using the same Savitzky-Golay machinery (also requires `signal`). Derivatives
emphasize subtle inflection points that are hard to see in raw reflectance.

The classic example is the **red edge**: the steep rise in reflectance between
the chlorophyll absorption in the red (~680 nm) and the bright NIR plateau
(~750 nm). Its inflection point -- the wavelength of steepest slope -- shows up
as a sharp peak in the **first derivative**, and its position shifts with
chlorophyll content and physiological stress. That is exactly what a derivative
exposes:

```{r, fig.height=3.5, fig.width=7, eval = requireNamespace("signal", quietly = TRUE)}
d1 = suppressMessages(deriv_spectra(one, order = 1))
d2 = suppressMessages(deriv_spectra(one, order = 2))

# The red-edge position: wavelength of the largest 1st-derivative value in the
# red/NIR transition (roughly 680-760 nm).
red_edge_win = bands(d1) >= 680 & bands(d1) <= 760
rep_nm       = bands(d1)[red_edge_win][which.max(value(d1)[1, red_edge_win])]

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(d1, main = "1st derivative", col = spec_pal[4],
     ylab = "dReflectance/dWavelength")
abline(h = 0, col = spec_grid, lty = 2)
abline(v = rep_nm, col = spec_pal[6], lwd = 1.5)
text(rep_nm, max(value(d1)[1, ]), labels = paste0("red edge\n~", rep_nm, " nm"),
     pos = 4, col = spec_pal[6], cex = 0.8)

plot(d2, main = "2nd derivative", col = spec_pal[5],
     ylab = "d2Reflectance/dWavelength2")
abline(h = 0, col = spec_grid, lty = 2)

par(oldpar)
```

`deriv_spectra()` (like `smooth_sgolay()` with a derivative order > 0) emits a
message reminding you that the output is **no longer reflectance/radiance** --
the y-axis units have changed. Pass `quiet = TRUE` once you know that. The
values still flow into any downstream function (`normalize()`, the two-band
spectral indices), so nothing stops you from processing them further; it is on
you to remember what the axis now means.

## Resampling to a common grid

Different instruments -- or the same instrument at different settings -- report
spectra on different band grids: different starting wavelength, spacing, or
count. Before you can compare or combine them, you must bring them onto a common
grid. `resample()` does this, but it is **not** naive linear interpolation: it
models each new band as a **Gaussian-weighted average of nearby original
bands**, using the **Full Width at Half Maximum (FWHM)** of the new band as the
Gaussian's spread. This mimics how a real detector integrates light over a small
window, rather than pretending the instrument measured an infinitely narrow
slice at each wavelength.

```{r, out.width="100%"}
new_bands = seq(400, 2400, 5)
fwhm      = make_fwhm(s, new_bands)

s_resamp  = resample(s, new_bands = new_bands, fwhm = fwhm)

plot(s, lwd = 0.5, col = spec_muted, main = "Original (grey) vs resampled (red)")
plot(s_resamp, add = TRUE, lwd = 0.25, col = spec_pal[6])
```

### Where the FWHM comes from

`resample()` needs one FWHM value per new band. `make_fwhm()` estimates a
sensible one from the spacing of the *original* bands, resampled (via the same
Gaussian model) onto the *new* grid, so a new band is never narrower than the
data that feeds it.

```{r}
length(fwhm) == length(new_bands)   # one value per new band
range(fwhm)
```

You can also supply your own FWHM (e.g. straight from the instrument's specs) --
a single number recycled across all new bands, or one value per new band.

### Resampling gotchas

**`make_fwhm()`'s default `k` bins the FWHM into just 3 clusters.** By default
`make_fwhm(spec, new_bands, k = 3)` runs a k-means step over the estimated FWHM
values and replaces each with its cluster's center. This keeps the number of
distinct resampling kernels small (cheaper to compute) when your new bands have
only a couple of natural resolutions, but it means the returned FWHM is **not** a
smooth function of wavelength even when the true one would be -- easy to mistake
the clustering artifact for a property of your data. The effect is invisible for
evenly spaced bands but obvious when they are not:

```{r}
# Dense in the VIS/NIR (2 nm), sparse in the SWIR (10 nm)
uneven_bands = c(seq(400, 1000, 2), seq(1005, 2400, 10))

fwhm_default = make_fwhm(s, uneven_bands)          # k = 3 (default)
fwhm_full    = make_fwhm(s, uneven_bands, k = 0)   # no clustering

length(unique(round(fwhm_default, 3)))  # only 3 distinct values...
length(unique(round(fwhm_full, 3)))     # ...vs the true band-to-band variation
```

Set `k = 0` for the uncollapsed, per-band FWHM estimate.

**Requested bands outside the original range are silently trimmed.**
`resample()` will not invent data beyond the edges of what you measured. Any
`new_bands` that fall (approximately) outside the original range are dropped,
with a message -- not extrapolated, and no error:

```{r}
too_wide = seq(330, 2525, 5)
s_trim   = resample(s, new_bands = too_wide, fwhm = make_fwhm(s, too_wide))

range(bands(s))       # what we actually measured
range(bands(s_trim))  # what we got back -- trimmed to fit
```

(Asking for bands *far* outside the original range -- tens of nanometers or more
-- does not just trim: the Gaussian weights in `make_fwhm()` underflow to zero
and it errors rather than returning nonsense. Either way, `resample()` will not
extrapolate.)

**Upsampling does not add real spectral detail.** Asking for a finer spacing
than your instrument records (e.g. 1 nm data to 0.1 nm) runs without error and
produces smoothly interpolated values, but it cannot recover detail the sensor
never captured. Treat upsampled data as a smoother *display* of the original
measurement, not as new information.

## Normalizing

Two leaves of the same species, measured moments apart, can differ by tens of
percent in overall brightness purely because of illumination geometry and
specular glare -- variation that swamps the biochemical signal you want.
**Vector normalization** removes it by dividing each spectrum by its own
magnitude (`sqrt(sum(value^2))`), so every spectrum has unit length and only its
*shape* remains.

```{r}
s_norm = normalize(s)
```

The payoff is clearest on the full dataset colored by species: before
normalizing, spectra of one species fan out vertically by brightness; after, the
within-species bands collapse and the species separate by shape alone.

```{r, fig.height=4, fig.width=8}
cols = species_cols(s)

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(s, lwd = 0.6, col = cols, main = "Raw (brightness scatter)")

plot(s_norm, lwd = 0.6, col = cols, main = "Vector normalized (shape only)")

par(oldpar)
```

The same tightening is visible as shrinking quantile envelopes. Here are the
95% per-band ribbons for one species before and after -- the normalized ribbon
is markedly narrower:

```{r, fig.height=3.5, fig.width=7}
sp        = names(s)[1]
one_sp    = s[names(s) == sp, ]
one_sp_n  = s_norm[names(s_norm) == sp, ]

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot_quantile(one_sp, total_prob = 0.95, col = spec_region_tints[2],
              border = spec_pal[2], main = paste(sp, "- raw"))
plot_quantile(one_sp_n, total_prob = 0.95, col = spec_region_tints[1],
              border = spec_pal[1], main = paste(sp, "- normalized"))

par(oldpar)
```

`normalize()` messages that the y values are no longer true reflectance (pass
`quiet = TRUE` to silence it) and stashes each spectrum's original magnitude in
metadata as `normalization_magnitude`, so the brightness you divided out is
recorded rather than lost.

## Recommended order

These steps compose, and the order matters. A sensible default pipeline is:

1. **`match_sensors()`** -- splice overlapping detector ranges so bands become
   strictly increasing. Everything downstream requires this.
2. **`smooth()`** -- suppress band-to-band noise on the native grid.
3. **`deriv_spectra()`** *(optional)* -- if your analysis works on derivatives
   (e.g. red-edge position), compute them here, remembering the y-axis is no
   longer reflectance.
4. **`resample()`** -- bring every spectrum onto the common band grid you will
   analyze on.
5. **`normalize()`** *(optional)* -- remove overall brightness so comparisons
   reflect spectral shape.

Every one of steps 2-5 requires **strictly increasing bands**, which is why
`match_sensors()` comes first. When in doubt, plot after each step: these
transforms are cheap, and a quick before/after view catches most mistakes before
they reach your analysis.
