---
title: "Combining and harmonizing datasets"
author: "Jose Eduardo Meireles, Anna K. Schweiger, and Jeannine Cavender-Bares"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Combining and harmonizing datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
source("_common.R")
library("spectrolab")
```

A field campaign is rarely one tidy file. It is a folder (or several) of
per-scan files, sometimes from more than one instrument, often on different
wavelength grids, and usually with several scans per leaf. The goal of this
vignette is to turn that mess into a single, analysis-ready `spectra` object:
put everything on a common band grid, stack the samples, and collapse replicate
scans into per-sample means -- without silently mixing things that should not
be mixed.

We use a small, real SVC (`.sig`) dataset shipped with the package: seven leaf
scans of a single species, *Acer platanoides*, read as **radiance**. The
filenames encode the sampling design -- a plot (`D2`, `F3`) and a canopy
position (`_B_`, `_M_`, `_T_` for bottom, middle, top) -- which gives us a clean
replicate structure to aggregate over.

```{r read-acer}
acer = read_spectra(system.file("extdata/Acer_example", package = "spectrolab"),
                    type               = "target_radiance",
                    exclude_if_matches = c("BAD", "WR"))

# Turn the filename design into metadata columns we can group by later.
parts = strsplit(names(acer), "_")
meta(acer, "plot")     = sapply(parts, `[`, 2)
meta(acer, "position") = sapply(parts, `[`, 4)

acer
```

## The problem

Two things get in the way of simply stacking a campaign together: the band
grids differ, and the *quantities* differ. Reading two more datasets that ship
with the package makes the point. The Acer scans are SVC radiance on a
1024-band grid; a second SVC dataset is **reflectance**; a PSR (`.sed`) dataset
is raw **digital numbers** on a 2151-band grid.

```{r heterogeneity}
serbin = read_spectra(system.file("extdata/svc_raw_and_overlap_matched_serbin/SVC_Files",
                                   package = "spectrolab"),
                      format = "sig")                        # SVC reflectance

psr    = read_spectra(system.file("extdata/psr_DN_brett", package = "spectrolab"),
                      format = "sed", type = "target_radiance")   # PSR digital numbers

# Same campaign spirit, three different band grids:
data.frame(
    dataset  = c("acer (radiance)", "serbin (reflectance)", "psr (DN)"),
    n_bands  = c(length(bands(acer)), length(bands(serbin)), length(bands(psr))),
    min_band = c(min(bands(acer)),    min(bands(serbin)),    min(bands(psr))),
    max_band = c(max(bands(acer)),    max(bands(serbin)),    max(bands(psr)))
)
```

The grids disagree in length (1024 vs. 2151) and, even where the lengths match,
in the actual wavelengths sampled. `combine()` refuses to guess: it stacks two
`spectra` only when their band labels are identical -- checked by both length
**and** `all.equal()`. That guard is exactly why harmonization comes first.
(The quantity mismatch -- radiance vs. reflectance vs. DN -- is a second, deeper
problem that no amount of resampling fixes; see
[An honest caveat](#an-honest-caveat).)

From here on we work entirely within the Acer radiance dataset, one species and
one quantity, so every operation is legitimate.

## Put everything on a common grid

Raw SVC data carries three overlapping detectors, so its bands are not strictly
increasing. `resample()` (like `normalize()`) requires strictly increasing
bands, so we first resolve the detector overlap with `match_sensors()` -- see
the dedicated *Matching sensor overlap* vignette for what that step does and how
to control it.

```{r match, message=FALSE}
acm = match_sensors(acer, splice_at = guess_splice_at(acer))
```

Now resample onto a target grid. `resample()` is not linear interpolation: it
models each output band as a Gaussian over the input bands, and `make_fwhm()`
picks a sensible full-width-half-maximum for each new band from the original
spacing. The key idea for harmonization is that **every dataset is resampled to
the same `new_bands`** -- that shared target is what makes them combinable.

```{r resample, message=FALSE}
new_bands = seq(400, 2400, by = 5)                       # one common target grid
acer_rs   = resample(acm, new_bands = new_bands, fwhm = make_fwhm(acm, new_bands))

acer_rs
```

## Combine

`combine()` stacks the samples of two `spectra` that share a band grid. If they
do not, it stops rather than producing a corrupt object -- the guard from the
problem section, seen here directly:

```{r combine-guard, error=TRUE}
# acm is on the native (997-band) grid; acer_rs is on the 5-nm grid.
combine(acm, acer_rs)
```

Once the grids agree, combining is trivial. Pretend the two plots were collected
as separate batches (different days, different folders); we split them, then
stack them back together. Metadata columns are carried through and merged by
name, so the reassembled object is identical to the original.

```{r combine}
batch_d2 = acer_rs[meta(acer_rs)$plot == "D2", ]
batch_f3 = acer_rs[meta(acer_rs)$plot == "F3", ]

campaign = combine(batch_d2, batch_f3)
campaign

# c() and rbind() are the idiomatic shorthands -- both just repeatedly
# apply combine(), so they preserve the spectra class and merge metadata.
identical(dim(c(batch_d2, batch_f3)), dim(campaign))
```

For more than two objects, `Reduce(combine, list(...))` (or `c()` / `rbind()`
with all of them) folds the whole list into one.

## Aggregate replicate scans

The campaign still holds replicate scans -- two per canopy position within plot
D2, plus one bottom scan from F3. `aggregate()` collapses those replicates into
per-group means. Grouping by canopy position turns seven scans into three
per-position mean spectra.

The subtlety is metadata. Averaging a numeric column is obvious, but text
columns (like `plot`) have no mean. Wrapping the metadata function in
`try_keep_txt()` tells `aggregate()` to fall back to the unique text values
instead of erroring or producing `NA` -- so the bottom-position mean correctly
records that it came from plots `D2, F3`.

```{r aggregate}
agg = aggregate(acer_rs,
                by       = meta(acer_rs)$position,
                FUN      = mean,
                FUN_meta = try_keep_txt(mean))

agg
meta(agg)
```

`split()` is the inverse operation: `split(acer_rs, meta(acer_rs)$position)`
returns a list of one `spectra` per group, which is handy when you want to
process groups independently before recombining them.

## An honest caveat

`spectrolab` deliberately does **not** track a `quantity` attribute on a
`spectra` object. Nothing in `combine()` or `aggregate()` will stop you from
stacking radiance on top of reflectance, or reflectance on top of digital
numbers -- the band-grid guard checks wavelengths, not physical units. That
responsibility is yours.

The worked example above is safe because it lives entirely inside one dataset:
Acer radiance, one species, one instrument. The three datasets from *The
problem* are **not** mutually combinable -- `acer` is radiance, `serbin` is
reflectance, `psr` is digital numbers -- and no resampling changes that.
Merging across instruments is an advanced case that requires matching the
quantity (convert to a common physical unit first) **and** the grid (resample
to shared `new_bands`) before `combine()` is meaningful.

## Before and after

The payoff: ragged per-scan inputs on the left, one harmonized object on the
right -- per-position mean spectra with 95% quantile ribbons, colored by canopy
position.

```{r payoff, fig.width=7.2, fig.height=3.8, message=FALSE}
gp = group_palette(acer_rs, by = "position")   # named colors, by position level

op = par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 0.5))

## Before: raw scans, overlapping detectors, colored by position
plot(acer, col = species_cols(acer, by = "position"), lwd = 0.7,
     cex.axis = 0.8, main = "Before: 7 raw scans")

## After: harmonized per-position means with quantile ribbons
plot(acer_rs, type = "n", cex.axis = 0.8, main = "After: harmonized means")
for(p in names(gp$pal)){
    sp = acer_rs[meta(acer_rs)$position == p, ]
    plot_quantile(sp, col = adjustcolor(gp$pal[[p]], 0.25),
                  border = FALSE, add = TRUE)
}
plot(agg, col = gp$pal[names(agg)], lwd = 2, add = TRUE)
legend("topright", legend = names(gp$pal), col = gp$pal,
       lwd = 2, bty = "n", cex = 0.8)

par(op)
```

From a folder of overlapping-detector scans on a vendor grid to a single object
on a clean 5-nm grid, with replicate scans averaged and their design metadata
intact -- ready for analysis.
