---
title: "Introduction to casper"
author: "Jose Eduardo Meireles"
date: "`r Sys.Date()`"
output:  rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Introduction to casper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The goal of `casper` is to provide a `spectra` class for `R` that exposes a standard interface and create a patafrom that allows other `R` packages to build on. The package will provide basic IO, plotting and conversion functionality, but that is about it. `casper` is implemented with ease of use in mind, but it should not slow you down.

## Installing and loading `casper`

The best way to get `casper` is to install it directly from the [_github repository_](www.github.com/meireles/casper). You will need the `devtools` package to do it.

```{r, eval=FALSE}
library("devtools")
install_github("meireles/casper")
```

Assuming that everything went smoothly, you should be able to load `casper` like any other package.

```{r}
library("casper")
```

## Reading spectra and creating a `spectra` object

### First, explore the example dataset `spec_matrix_example`

`casper` comes with limited IO capabilities. To illustrate how to create a `spectra` object, we will use an example dataset called `spec_matrix_example`. Samples are in rows, and wavelengths in columns, and the first column is the sample name (in this case, a species name). The column names match wavelength labels. I tried to format `spec_matrix_example` as a typical spectral dataset in `csv` format.

```{r}
# Example spectral dataset in matrix format.
spec_matrix_example[1:4, 1:3]

# Note that this is NOT a spectra object. You can verify this by either asking what class `spec_example` is or using casper's `is_spectra()` function.

class(spec_matrix_example)
is_spectra(spec_matrix_example)
```

### Constructing a `spectra` object

The `spectra` class holds the essential information used in spectral dataset: reflectance, wavelengths, etc. The class has a bunch of requirements in terms of both format and values. For instance, relative reflectance must be between 0 and 1.

If your data is in a matrix with the same format as `spec_matrix_example` (check previous section for details), you can construct a `spectra` object by calling the `as.spectra()` function.

```{r}
# Make a spectra object if you have a matrix in the right format
spec = casper::as.spectra(spec_matrix_example)

# Did it work?
is_spectra(spec)
```

Alternativelly, you can create a `spectra` object using the more flexible `spectra()` constructor, which takes three arguments: (1) a reflectance matrix, (2) a vector of wavelengths and (3) the sample names.

```{r}
# (1) Create a reflectance matrix.
#     In this case, by removing the species column
rf = spec_matrix_example[ , -1 ]

# Check the result
rf[1:4, 1:3]

# (2) Create a vector with wavelength labels that match
#     the reflectance matrix columns.
wl = colnames(rf)

# Check the result
wl[1:4]

# (3) Create a vector with sample labels that match
#     the reflectance matrix rows.
#     In this case, use the first colum of spec_matrix_example
sn = spec_matrix_example[ , 1] 

# Check the result
sn[1:4]

# Finally, construct the spectra object using the `spectra` constructor
spec = spectra(reflectance = rf, wavelengths = wl, names = sn)

# And hopefully this worked fine
is_spectra(spec)
```

### Converting a `spectra` object into a matrix

It is possible to convert a `spectra` object to a matrix format, using the `as.matrix()` function. `casper` will (1) place wavelengths in columns, assigning wavelength labels to `colnames`, and (2) samples in rows, assigning sample names to `rownames`. Since `R` imposes strict on column name formats and sometimes on row names, `as.matrix()` will try to fix potential dimname issues if `fix_names != "none"`.

```{r, error=TRUE}
# Make a matrix from a `spectra` object
spec_as_mat = as.matrix(spec, fix_names = "none")
spec_as_mat[1:4, 1:3]
```


# Exploring a `spectra` object

`casper` exposes a few ways to plot and query spectral data in `spectra` format. TODO ... 

## Plotting

The workhorse function for plotting `spectra` is `plot()`. It will jointly plot each spectrum in the `spectra` object. You should be able to pass the usual plot arguments to it, such as `col`, `ylab`, etc.

You can also plot the quantile of a `spectra` object with `plot_quantile()`. It's second argument, `total_prob`, is the total "mass" that the quantile encompasses. For instance, a `total_prob = 0.95` covers 95% of the variation in the `spectra` object; i.e. it is the `0.025 to 0.975` quantile. The quantile plot can stand alone or be added to a current plot if `add = TRUE`.

Last but not least, you can also shade spectral regions with the `plot_spec_regions()` function. `casper` provides a `default_spec_regions()` matrix as an example, but you obviously can customize it for your needs.

```{r, fig.height=2.5, fig.width=8}
par(mfrow = c(1, 3))

# Simple spectra plot
plot(spec, lwd = 0.75, lty = 1, col = "grey25", main = "All Spectra")

# Stand along quantile plot
plot_quantile(spec, total_prob = 0.8, main = "80% spectral quantile", 
              col = rgb(1, 0, 0, 0.5), lwd = 0.5, border = TRUE)

# Combined quantile and individual spctra plot
# With an added bonus of shading 4 spectral regions
plot(spec, lwd = 0.25, lty = 1, col = "grey50", "Spectra, quantile and regions")
plot_quantile(spec, total_prob = 0.8, 
              col = rgb(1, 0, 0, 0.25), add = TRUE, border = FALSE)
plot_spec_regions(spec, regions = default_spec_regions(), add = TRUE)
```

## Querying

`casper` lets you query the `spectra` object and get summary infomation. You can easly get sample names with `names()` and wavelength labels with `wavelengths()`. It is also possible to recover the 

```{r}
# Get the vector of all sample names
# Note that duplicate sample names are permitted
n = names(spec)
n[1:5]

# Or get the vector of wavelengths
w = wavelengths(spec)
w[1:5]

# You can also get the dimensions of your `spectra` object
dim(spec)
```

If you really need the raw reclectance, you can retrieve it with the `reflectance()` function.

## Subsetting spectra

You can subset the `spectra` using a notation *similar* to the `[ i , j]` function used in matrices and data.frames. The first argument in `[ i, ]` matches *sample names*, whereas the second argument `[ , j ]` matches the *wavelength names*. Here are some examples of how `[` works in `specrta`:

  - `x[ 1:3 , ]` will keep the first three samples of `x`. I.e. `1:3` are indexes.
  - `x[ "sp_1" , ]` keeps **all** entries in `x` where sample names match `"sp_1"`
  - `x[  , 800:900 ]` will keep wavelengths between `800` and `900`.
  - `x[  , 1:5 ] ` will **fail**!. *wavelengths __cannot__ be subset by index!*

```{r, fig.height=2.5, fig.width=6, fig.align="center"}
# Subset spectra to all entries where sample_name matches "species_8"
spec_sp8 = spec[ "species_8", ]

# Check the results
dim(spec_sp8)

# Plotting the seubset result should work just fine
par(mfrow = c(1, 2), cex.main = 0.8, cex.axis = 0.6, cex.lab = 0.8)

plot(spec_sp8, col = "red", main = "Species 8 spectra")
plot_quantile(spec, total_prob = 1.0, add = TRUE,  col = rgb(0.2, 0.2, 0.2, 0.2), border = FALSE)
plot_spec_regions(spec_sp8, default_spec_regions(), col = rgb(1, 0.5, 0, 0.1), add = TRUE)

# And maybe further subset to the visible wavelengths only
spec_sp8 = spec_sp8[ , 400:700 ]

# This subset should still plot just fine
plot(spec_sp8, col = "red", main = "Visible spectra for species 8")
plot_quantile(spec, total_prob = 1.0, add = TRUE,  col = rgb(0.2, 0.2, 0.2, 0.2), border = FALSE)
```

Note that you can (1) subset samples using indexes and (2) use character or numeric to subset wavelengths. As said before, you cannot use indexes to subset wavelengths though.

```{r, error=TRUE}
# Subset samples by index should work. It is also OK to subset by wavelength 
# using numerics or characters.
reflectance(spec_sp8[ 1 , "405"]) == reflectance(spec_sp8[ 1 , 405])

#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
# But remember that you CANNOT use indexes to subset wavelengths!
#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#

# Something that is obvioulsy an index, like using 2 instead of 401 (the 2nd band), will fail.
spec_sp8[ , 2 ]

# However, `casper` canot detect if you meant to get the two last bands when
# you use 2000:2001. It will assume that you wanted wavelengths "2000" and "2001"
# Bottomline, be very careful not to use indexes to subset wavelengths!
```

## Manipulating samples and wavelength labels

You may want to edit certain simple attributes of `spectra`, such as make all sample names uppercase. This is easlily attainable in `casper`:

```{r}
spec_new = spec

# Replace names with an uppercase version
names(spec_new) = toupper(names(spec_new))

# Check the results
names(spec_new)[1:5]
```

You may want to fiddle with the reflectance itself. This is easy to do, but there are some constraints. For example, `casper` will not allow you to have negative reflectance values or values greater than 1.

```{r,  fig.height=3, fig.width=4, fig.align="center"}
# Scale reflectance by 0.75
# spec_new[] = reflectance(spec_new) * 0.75
spec_new = spec_new * 0.75

# Plot the results
plot(spec, col = "blue", lwd = 0.75, cex.axis = 0.75)
plot(spec_new, col = "orange", lwd = 0.75, add = TRUE)
```

However, `casper` will throw an error if you try to perform an illegal operation to reflectance, for instance 

```{r, error=TRUE}
# Trying to add 1.0 to all reflectance values will fail.
spec_new[] = reflectance(spec_new) + 1.0
```
