---
title: "Introduction to spectrolab"
author: "Jose Eduardo Meireles, Anna K. Schweiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spectrolab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The goal of `spectrolab` is to provide essential functions to process data from portable spectrometers and a common interface for spectra that other packages can build on. Currently `spectrolab` has been only tested for SVC data (Spectral Vista Inc.). `spectrolab` provides methods to read spectral data files, to perform jump correction, vector normalization, smoothing at the sensor overlap regions, and to filter out bad measurements. `spectrolab` introduces the S3 class `spectra`, which makes it easy to plot and manipulate data and to access additional metadata.  

## Installing and loading `spectrolab`

The best way to get `spectrolab` is to install it directly from the [_github repository_](www.github.com/annakat/spectrolab). You will need the `devtools` package to do it.

```{r, eval=FALSE}
library("devtools")
install_github("annakat/spectrolab")
```

Assuming that everything went smoothly, you should be able to load `spectrolab` like any other package.

```{r, eval=FALSE}
library("spectrolab")
```

## Reading spectra: example with `.sig` files

You can read raw spectra files using the function `read_spectra()`. As of now, only files in the `.sig` (Spectra Vista) can be read direcly, but other formats are in the works.
You can pass a vector of filenames to `read_spectra()`, but it is usually easier to pass the path for the folder where all your data is. 

```{r, eval=FALSE}

# dir_path is the directory where our example datasets live
dir_path <- system.file("extdata", "Acer_example", package = "spectrolab")

# Read .sig files
acer_spectra <- read_spectra(path = dir_path, format ="sig")

# Note that `acer_spectra` is a `spectra` object. You can ensure that is true
# using spectrolab's `is_spectra()` function.

is_spectra(acer_spectra)
```

You can inspect the newly read data `acer_spectra` in several ways. You may want to know how many spectra and how many bands are in there or may want to plot them.

```{r, eval=FALSE}
# Simply print the object
acer_spectra

# Get a vector with the dataset dimension
dim(acer_spectra)

# and plot the spectra
plot(acer_spectra)
```

You notice that `Acer_example` contains white reference measurement (the horizontal line around 1.0 reflectance) and one measurement obviously bad measurement (the one with the dip at the NIR shoulder). During the measurement campaign those two measurement were flagged with "_WR" and "_BAD". With  `read_spectra()` it's easy to exclude flagged measurements. 

```{r, eval=FALSE}

# Use `exclude_if_matches` to excluded flagged files
acer_spectra <- read_spectra(path = dir_path, format ="sig",   exclude_if_matches = c("BAD","WR"))

# and check result
plot(acer_spectra)
```
Flagging "unusual" files during a measurement campaign is a very good idea, as it clearly helps you to quickly sort through your files when it comes to processing. However, you can also exclude "unusual" files lateron. We show you some build in options and how to filter out spectra based on your own criteria in section `Excluding "unusual" spectra` below. 

### Create spectra from a matrix

If you already have your spectra in a matrix, you can use the function `as.spectra()` to convert it to a `spectra` object. The matrix **must** be in a very specific format, so **read carefully!** Samples must be in rows and wavelengths in columns. The first column must be a sample label or filename. In case you are as terrified as Ms. A of loosing metadata for 100's of thousands of spectra, consider including "all" crucial sample information in the filename (`species_plotID_plantID`). Column names must be the wavelength labels. We included an example of a matrix like that named `spec_matrix_example`.

```{r, eval=FALSE}
spec <- spec_matrix_example

# Check out the format of the matrix
spec[1:3, 1:4]

# To convert it to spectra, simply run
spec_from_matrix <- as.spectra(spec)

# and again you can plot it to make sure everything worked okay
plot(spec_from_matrix)
```

### Converting a `spectra` object into a matrix

It is also possible to convert a `spectra` object to a matrix format, using the `as.matrix()` function. `spectrolab` will (1) place wavelengths in columns, assigning wavelength labels to `colnames`, and (2) samples in rows, assigning sample names to `rownames`. Since `R` imposes strict on column name formats and sometimes on row names, `as.matrix()` will try to fix potential dimname issues if `fix_names != "none"`.

```{r, error=TRUE}
# Make a matrix from a `spectra` object
spec_as_mat = as.matrix(spec_from_matrix, fix_names = "none")
spec_as_mat[1:4, 1:3]
```


## Plotting

The workhorse function for plotting `spectra` is `plot()`. It will jointly plot each spectrum in the `spectra` object. You should be able to pass the usual plot arguments to it, such as `col`, `ylab`, etc.

You can also plot the quantile of a `spectra` object with `plot_quantile()`. It's second argument, `total_prob`, is the total "mass" that the quantile encompasses. For instance, a `total_prob = 0.95` covers 95% of the variation in the `spectra` object; i.e. it is the `0.025 to 0.975` quantile. The quantile plot can stand alone or be added to a current plot if `add = TRUE`.

Last but not least, you can also shade spectral regions with the `plot_spec_regions()` function. `spectrolab` provides a `default_spec_regions()` matrix as an example, but you obviously can customize it for your needs.

```{r, fig.height=2.5, fig.width=8, error=TRUE}
# Simple spectra plot
par(mfrow = c(1, 3))
plot(spec_from_matrix, lwd = 0.75, lty = 1, col = "grey25", main = "All Spectra")

# Stand along quantile plot
plot_quantile(spec_from_matrix, total_prob = 0.8, col = rgb(1, 0, 0, 0.5), lwd = 0.5, border = TRUE)
title("80% spectral quantile")

# Combined quantile and individual spctra plot
# With an added bonus of shading 4 spectral regions
plot(spec_from_matrix, lwd = 0.25, lty = 1, col = "grey50", main="Spectra, quantile and regions")
plot_quantile(spec_from_matrix, total_prob = 0.8, col = rgb(1, 0, 0, 0.25), border = FALSE, add=TRUE)
plot_spec_regions(spec_from_matrix, regions = default_spec_regions(), add = TRUE)
```

## Querying

`spectrolab` lets you query the `spectra` object and get summary infomation. You can easly get sample names with `names()` and wavelength labels with `wavelengths()`. It is also possible to recover the 

```{r, eval=F}
# Get the vector of all sample names
# Note that duplicate sample names are permitted
n = names(spec_from_matrix)
n[1:5]

# Or get the vector of wavelengths
w = wavelengths(spec_from_matrix)
w[1:5]

# You can also get the dimensions of your `spectra` object
dim(spec_from_matrix)
```

If you need the reclectance, you can retrieve it with the `reflectance()` function.

## Subsetting spectra

You can subset the `spectra` using a notation *similar* to the `[ i , j]` function used in matrices and data.frames. The first argument in `[ i, ]` matches *sample names*, whereas the second argument `[ , j ]` matches the *wavelength names*. Here are some examples of how `[` works in `specrta`:

  - `x[ 1:3 , ]` will keep the first three samples of `x`. I.e. `1:3` are indexes.
  - `x[ "sp_1" , ]` keeps **all** entries in `x` where sample names match `"sp_1"`
  - `x[  , 800:900 ]` will keep wavelengths between `800` and `900`.
  - `x[  , 1:5 ] ` will **fail**!. *wavelengths __cannot__ be subset by index!*

For example, you can exclude the noisy regions at the beginning and end of the spectrum and subset the data to specific entries.

```{r, fig.height=2.5, fig.width=6, fig.align="center", eval=F}

### Subset wavelength regions 
spec_sub <-  spec_from_matrix[ , 400:700]

### Check the result
plot(spec_sub)

# Subset spectra to all entries where sample_name matches "species_8" and wavelength regions 
spec_sp8 = spec_from_matrix[ "species_8", ]

# Check the results
dim(spec_sp8)

# Plotting the subset result should work just fine
plot(spec_sp8, col = "red", main = "Species 8 spectra")
plot_quantile(spec_sp8, total_prob = 0.75, add = TRUE,  col = rgb(0.2, 0.2, 0.2, 0.2), border = TRUE)
plot_spec_regions(spec_sp8, default_spec_regions(), col = rgb(1, 0.5, 0, 0.1), add = TRUE)
```

Note that you can (1) subset samples using indexes and (2) use character or numeric to subset wavelengths. As said before, you cannot use indexes to subset wavelengths though.

```{r, error=TRUE}
# Subset samples by index should work. It is also okay to subset by wavelength using numerics or characters.
reflectance(spec_sp8[ 1 , "405"]) == reflectance(spec_sp8[ 1 , 405])

#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
# But remember that you CANNOT use indexes to subset wavelengths!
#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#

# Something that is obvioulsy an index, like using 2 instead of 401 (the 2nd band), will fail.
spec_sp8[ , 2]

# However, `spectrolab` canot detect if you meant to get the two last bands when
# you use 2000:2001. It will assume that you wanted wavelengths "2000" and "2001"
# Bottomline, be very careful not to use indexes to subset wavelengths!
```

## Manipulating samples and wavelength labels

You may want to edit certain simple attributes of `spectra`, such as make all sample names uppercase. This is easlily attainable in `spectrolab`:

```{r, eval=F}
spec_new <- spec_from_matrix

# Replace names with an uppercase version
names(spec_new) <- toupper(names(spec_new))

# Check the results
names(spec_new)[1:5]
```

You may want to fiddle with the reflectance itself. This is easy to do.

```{r,  fig.height=3, fig.width=4, fig.align="center", eval=F}
# Scale reflectance by 0.75
# spec_new[] = reflectance(spec_new) * 0.75
spec_new <- spec_new * 0.75

# Plot the results
plot(spec_from_matrix, col = "blue", lwd = 0.75, cex.axis = 0.75)
plot(spec_new, col = "orange", lwd = 0.75, add = TRUE)
```

### Jump correction

Some instruments allow you to preserve the sensor overlap regions. This results in duplicated wavelenghts and a "jump" between the VIS/NIR and NIR/SWIR regions. The `jump corrector` removes the overlap regions between the VIS/NIR and NIR/SWIR regions, respectively, and corrects the spectra to allow a smooth transition, while keeping the information as original as possible. In principle, the algorithm searches for the wavelengths of closest approach in both overlap regions, cuts the spectrum into three pieces, removes the overlap and finds a multiplier to paste the junks back together. No additional smoothing is applied. As a convention, the NIR region is left unchanged.  

```{r, eval=F}
# Read Acer example spectra 
acer_spectra <- read_spectra(path = dir_path, format ="sig",   exclude_if_matches = c("BAD","WR"))

# and check result
plot(acer_spectra)

# Jump correction 
acer_juco <- jump_corr(acer_spectra)

# Looks better!
plot(acer_juco)
```

### Smoothing at overlap regions

Depending on your analysis, it can be a good idea to smooth remaining noise at the sensor overlap regions. `spectrolab` comes with two implemented smoothing functions: `smoo_visnir` uses a moving average between 970 - 1025 nm. `smoo_nirswir` uses linear interpolation between 1903 - 1917 nm. Note: Ms. A does not do this when predicting traits from spectra. 

```{r, eval=FALSE}
### Smooth only VIS/NIR or NIR/SWIR
Acer_smoo1 <- smoo.visnir.svc(Acer_juco)
Acer_smoo2 <- smoo.nirswir.svc(Acer_juco)

### Smooth both regions
Acer_smoo <- smoo.nirswir.svc(Acer_smoo1)
### Same result
Acer_smoo <- smoo.visnir.svc(Acer_smoo2)
```

## Excluding "unusual" spectra

The `spectrolab` package comes with a couple of functions to exclude bad spectra. However, thus far the functions only accept data.frames, such as Acer_juco, the result from the initial jump correction. At this point, a typical workflow would involve inspecting the raw spectra with `plot_rawspec`, removing bad spectra, and plotting again, until the output looks fine. Then a `spectra` object can be constructed with `as.spectra`.

```{r, eval=FALSE}
### Some common examples, see `excl.` for more.
### Exclude spectra with reflectances at the 'NIR shoulder' @ 761 nm <0.3 or >0.65 
Acer_excl <- exclhilo_spec(Acer_juco)

### Exclude spectra with high reflectances in VIS: @ 450 nm >0.2 
Acer_excl <- exclhivis_spec(Acer_juco)

### Exclude spectra with dips in NIR: @800 - @770 >0.02 ###
Acer_excl <- excldipnir_spec(Acer_juco)

### Create spectra object and plot
Acer_corr <- as.spectra (Acer_excl)
plot(Acer_corr)
```




