---
title: "Introduction to spectrolab"
author: "Jose Eduardo Meireles, Anna K. Schweiger, and Jeannine Cavender-Bares"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to spectrolab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
source("_common.R")
library("spectrolab")
```

`spectrolab` reads, processes, and visualizes data from portable
spectroradiometers. Its foundation is the `spectra` S3 class: a lightweight,
value-semantic container for reflectance or radiance spectra that stays
dependency-light and plays well with base R. This vignette is the quickstart
hub -- read data in, inspect and subset it, get it back out, and compute a
first result -- with pointers to the focused vignettes for everything else.

The figure below plots a labeled, multi-species dataset so you can see where
this is going: 50 samples across 10 groups, colored by group, with the
visible, near-infrared, and short-wave infrared regions shaded behind the data.

```{r hook, fig.height=4.2, fig.width=7, echo=TRUE}
# SIMULATED illustrative data: 50 samples x 2101 bands (400-2500 nm),
# grouped into 10 "species". Not real measurements.
s = as_spectra(spec_matrix_example, name_idx = 1)

plot(s, col = species_cols(s), lwd = 0.8, cex.axis = 0.8,
     main = "Simulated illustrative spectra, colored by species")
plot_regions(s, col = spec_region_tints, add = TRUE)
```

## Reading spectra

`read_spectra()` reads raw spectroradiometer files -- SVC `.sig`, Spectral
Evolution `.sed`, and ASD `.asd`. Pass it either a vector of file names or,
more commonly, the path to a folder of files. `spectrolab` guesses the format,
but you can force it with `format`. It will not descend into nested folders or
mix file types in one call.

The package ships a folder of real SVC `.sig` scans of *Acer platanoides*
(Norway maple) leaves. These are **radiance**, not reflectance -- keep that in
mind when interpreting values and indices.

```{r read-acer}
dir_path = system.file("extdata/Acer_example", package = "spectrolab")

# Field crews often flag files: "_WR" for white references, "_BAD" for bad
# scans. `exclude_if_matches` drops any file whose name contains these tokens.
acer = read_spectra(dir_path, exclude_if_matches = c("BAD", "WR"))

acer
```

Set `extract_metadata = TRUE` to also pull each file's header (instrument,
integration times, GPS, timestamps, ...) into the object's metadata.

```{r read-meta, message=FALSE}
acer_m = read_spectra(dir_path,
                      exclude_if_matches = c("BAD", "WR"),
                      extract_metadata   = TRUE)

# A few header fields for the first three scans
meta(acer_m)[1:3, c("instrument", "integration1", "units")]
```

Multi-detector instruments (the SVC HR1024 uses separate Si/InGaAs sensors)
also record where detectors meet. That splice provenance rides along as the
`sensor_info(x)` attribute -- not a `spectra` slot -- and is what lets
`match_sensors()` find the splice points for you. See
[Matching sensor overlap](match_sensors.html).

### From a matrix or data.frame

If your spectra are already in a matrix or data.frame -- for example, read from
a CSV -- convert them with `as_spectra()`. Samples must be in rows and bands in
columns, with **numeric band labels as the column headers**. Name the column
holding sample names with `name_idx`, and any other non-band columns with
`meta_idxs`.

```{r as-spectra}
csv_path = system.file("extdata/spec_matrix_meta.csv", package = "spectrolab")

# check.names = FALSE keeps numeric band headers intact; otherwise R prepends
# an "X" (e.g. "X650"), which breaks the band labels.
d = read.csv(csv_path, check.names = FALSE)

# Column 3 holds sample names; columns 1-2 are metadata.
achillea = as_spectra(d, name_idx = 3, meta_idxs = c(1, 2))
achillea
```

## Inspecting and subsetting

Print a `spectra`, or query its parts: `dim()`, `nrow()`, `names()` (sample
names), `bands()` (band labels), `value()` (the matrix), and `meta()`.

```{r inspect}
dim(acer)          # samples x bands
nrow(acer)         # samples
head(names(acer))  # file names became sample names
range(bands(acer)) # first and last wavelength
```

Subset with `[i, j]`, but note the twist: `i` matches **sample names or
index**, while `j` matches **band labels, not column position**. So `x[, 600]`
is the 600 nm band, and `x[, 1:5]` would (incorrectly) ask for bands *labeled*
1 through 5.

```{r subset}
# Samples by index; bands by label
vis = achillea[, 400:700]

# All samples whose name matches "ACHMI_7"
one = achillea["ACHMI_7", ]

# For non-integer band grids, select a range with bands(x, min, max)
acer_trim = acer[, bands(acer, 400, 2400)]
```

`subset_by()` keeps only groups with between `n_min` and `n_max` members --
handy for dropping under-replicated groups before summarizing.

```{r subset-by}
# Keep only species with at least 3 samples (all 10 qualify here)
common = subset_by(s, by = names(s), n_min = 3, n_max = Inf)
dim(common)
```

**A quick QA beat.** Plotting a quantile envelope over the raw lines makes
outliers pop out -- a scan that strays outside the bulk of the data is easy to
spot visually before it contaminates a summary.

```{r qa-envelope, fig.height=4.2, fig.width=7}
plot(s, col = spec_muted, lwd = 0.5, cex.axis = 0.8,
     main = "Raw spectra with an 80% quantile envelope")
plot_quantile(s, total_prob = 0.8, col = spec_seq(3)[2], border = FALSE, add = TRUE)
```

For large datasets, `plot_interactive()` opens a Shiny app to scroll and zoom
through spectra one screenful at a time. It is interactive, so it cannot render
in a static vignette; run it in an R session:

```{r interactive, eval=FALSE}
plot_interactive(s)
```

## Getting data out

Three ways to hand spectra to other tools:

```{r out}
# 1. A plain matrix: bands in columns, samples in rownames. Drops metadata.
m = as.matrix(achillea, fix_names = "none")
m[1:3, 1:3]

# 2. A data.frame, keeping metadata columns by default.
df = as.data.frame(achillea, fix_names = "none")
df[1:3, 1:4]

# 3. Long/tidy form: one row per sample-and-band (ideal for ggplot2).
lng = to_long(s)
head(lng, 3)
```

`as_tibble()` and `autoplot()` build on `to_long()`; see
[Visualizing spectral variation](visualizing_spectral_variation.html).

## A first result: spectral indices

Two-band normalized-difference indices are one line each. The built-in
`spectral_index` list carries ready-to-use functions -- call them directly on a
`spectra` object, which returns one value per sample.

```{r indices}
ndvi = spectral_index$ndvi(s)   # (R800 - R680) / (R800 + R680)
pri  = spectral_index$pri(s)    # (R570 - R531) / (R570 + R531)

summary(ndvi)
```

Build your own with `make_spectral_index(band1, band2)`, which returns a
function you can reuse. Here is a green-band NDVI variant:

```{r custom-index}
gndvi = make_spectral_index(800, 550)
head(round(gndvi(s), 3))
```

The index engine snaps each requested wavelength to the **nearest available
band**, but only within `tolerance` (default 1, in band units). If no band
falls within `tolerance` it errors rather than silently using a distant band --
so widen `tolerance` for coarse or resampled grids. Indices also require
strictly increasing bands: splice sensor overlaps first (see below).

For the simulated data, NDVI clusters high (dense-canopy-like reflectance) as
expected; PRI is centered near zero, consistent with its narrow physiological
range.

## Good to know

A few intentional behaviors that can still surprise you.

**Duplicate sample names *and* duplicate band labels are both allowed.**
Several scans of one individual share a name; raw SVC files legitimately carry
the same wavelength on both sides of a sensor overlap before splicing.
Subsetting a duplicated label returns *all* matches (and says so). But most
transforms (`smooth()`, `normalize()`, `resample()`, spectral indices) require
strictly increasing bands and will error on raw, unspliced overlaps -- run
`match_sensors()` first.

**`s1 + s2` is arithmetic; `combine()` stacks.** The overloaded operators
(`+ - * / ^`) act element-wise on two `spectra` of identical dimensions and
bands, and keep names/metadata only when both sides agree (otherwise they clear
with a warning). To row-bind two datasets -- bands checked for an exact match,
metadata columns merged by name -- use `combine()` (or its shorthands `c()` /
`rbind()`).

**Comparison operators return a bare matrix.** `s1 > s2`, `s1 == s2`, and
friends return a logical samples-by-bands matrix, not a `spectra` -- a boolean
"spectrum" is not meaningful. Only the arithmetic operators return `spectra`.

## Where to go next

* [Matching sensor overlap](match_sensors.html) -- splice multi-detector data.
* [Cleaning & standardizing](clean_and_standardize.html) -- smooth, trim, normalize.
* [Combining & harmonizing](combine_and_harmonize.html) -- stack and align datasets.
* [Estimating leaf traits with PLSR](leaf_traits_plsr.html) -- a modeling workflow.
* [Visualizing spectral variation](visualizing_spectral_variation.html) -- tidy data and `ggplot2`.
