---
title: "Introduction to spectrolab"
author: "Jose Eduardo Meireles, Anna K. Schweiger, and Jeannine Cavender-Bares"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to spectrolab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse    = TRUE,
    comment     = "#>",
    fig.align   = "center",
    fig.retina  = 2,
    dpi         = 150
)
```

`spectrolab` provides methods to read, process, and visualize data from portable
spectroradiometers. The package establishes a common interface for spectra by
introducing a `spectra` S3 class -- a lightweight container that plays well
with base R while staying dependency-light.

`spectrolab` packs a lot of functionality:

* Read spectra from raw spectral files (SVC, ASD, PSR/Spectral Evolution) or matrices
* Access, aggregate, subset, split or combine spectra
* Seamlessly link and manipulate metadata (such as chemistry or instrument metadata)
* Plot spectra or spectral quantiles, shade spectral regions (e.g. SWIR)
* Interactively scroll through and zoom in on spectra
* Smooth, resample, normalize, differentiate, and remove the continuum from spectra
* Splice sensor overlaps and compute two-band vegetation indices (e.g. NDVI, PRI)
* Reshape spectra into tidy/long data and plot them with `ggplot2`

This vignette covers the foundations: reading data in, inspecting and
subsetting a `spectra` object, plotting, and converting to/from matrices and
data.frames. Everything past that -- sensor-overlap splicing, resampling,
smoothing/derivatives/continuum removal, spectral indices, and the tidy
bridge -- has its own focused vignette; see [Where to go next](#where-to-go-next).

The source code can be found on our [GitHub repository](https://github.com/meireles/spectrolab).
Please report any bugs and ask us your questions through the [issue tracker](https://github.com/meireles/spectrolab/issues).


## Installing and loading `spectrolab`

The latest stable version of `spectrolab` is on [CRAN](https://cran.r-project.org/package=spectrolab). Install it with:

```{r, eval=FALSE}
install.packages("spectrolab")
```

You can also install it directly from GitHub using the `devtools` package.

```{r, eval=FALSE}
library("devtools")
install_github("meireles/spectrolab")
```

Assuming that everything went smoothly, you should be able to load `spectrolab` like any other package.

```{r, echo=TRUE, message=FALSE}
library("spectrolab")
```

## Reading spectra

There are two ways to get spectra into R using `spectrolab`:

  1. Reading spectra from raw data files (formats: SVC's `sig`, Spectral Evolution's `sed`, and ASD's `asd`).
  2. Converting a matrix or data.frame to `spectra`.

### Reading spectra from raw data files: Example with SVC's `.sig`

Use the function `read_spectra()` to read your spectroradiometer's files. You can pass a vector of file names to `read_spectra()` but it is usually easier to pass the path to the folder where your data are. Note that `spectrolab` cannot read nested folders or read a mix of file types at once (i.e. having `.sig` and `.sed` files in the same folder will produce an error).

```{r, eval=TRUE}
# `dir_path` is the directory where our example datasets live
dir_path = system.file("extdata/Acer_example", package = "spectrolab")

# Read spectra from .sig files inside the "extdata/Acer_example" folder
acer_spectra = read_spectra(path = dir_path)
```

`spectrolab` guesses the file format automatically (but you can provide the format using the `format` argument if needed). The package reads the target's relative reflectance -- the ratio between the target's radiance and the reference's radiance -- by default. However, you can read the target's or reference's radiances using the argument `type`.

```{r, eval=TRUE}
# Reading the target's radiance
acer_spectra_rad = read_spectra(path = dir_path, format = "sig", type = "target_radiance")

# And the white reference's radiance
acer_white_ref = read_spectra(path = dir_path, type = "reference_radiance")
```

You can avoid importing undesirable spectra if those were flagged in the field. For example, we usually add the suffixes "_WR" to denote white reference and "_BAD" to denote bad measurements, so we can pass those flags to the argument `exclude_if_matches` in `read_spectra()`.

```{r, eval=TRUE}
# Use the `exclude_if_matches` argument to excluded flagged files
acer_spectra = read_spectra(path = dir_path, exclude_if_matches = c("BAD","WR"))
```

Finally, `spectrolab` lets you read the metadata from `sig` and `sed` files if you need to. Simply set `extract_metadata` to `TRUE`.

```{r, eval=TRUE, message=FALSE}
# Use the `exclude_if_matches` argument to excluded flagged files
acer_spectra_with_meta = read_spectra(path = dir_path,
                                      exclude_if_matches = c("BAD","WR"),
                                      extract_metadata = TRUE)

# Here are fields 2 to 6 of the metadata for the first 3 scans.
# More on the `meta` function later.

meta(acer_spectra_with_meta)[1:3, 2:6]
```

> **Where did that come from?** Instruments that use more than one detector
> (e.g. SVC's HR1024, which uses separate Si/InGaAs sensors for VIS, SWIR1 and
> SWIR2) also record splice provenance at read time -- the wavelengths where
> detectors meet and, for some vendors, the matching factors applied. That
> information rides along as the `sensor_info(x)` attribute rather than as a
> `spectra` slot, and it is what lets `match_sensors()` know how to re-splice
> the data without you having to remember the boundaries; see the
> [Matching sensor overlap](match_sensors.html) vignette.

### Create spectra from a matrix or data.frame

If you already have your spectra in a matrix or data frame (e.g. when you read your data from a .csv file), you can use the function `as_spectra()` to convert it to a `spectra` object. The matrix **must** have samples in rows and bands in columns. The header of the bands columns must be (numeric) band labels. You also should declare which column has the sample names (which are mandatory) using the `name_idx` argument. If other columns are present (other than sample name and values), their indices must be passed to `as_spectra` as the `meta_idxs` argument.

Here is an example using a dataset matrix named `spec_matrix_meta.csv` provided by the package.

```{r, eval=TRUE}
dir_path = system.file("extdata/spec_matrix_meta.csv", package = "spectrolab")

# Read data from the CSV file. If you don't use `check.names` = FALSE when reading
# the csv, R will usually add a letter to the column names (e.g. 'X650') which will
# cause problems when converting the matrix to spectra.
spec_csv = read.csv(dir_path, check.names = FALSE)

# The sample names are in column 3. Columns 1 and 2 are metadata
achillea_spec = as_spectra(spec_csv, name_idx = 3, meta_idxs = c(1,2) )

# And now you have a spectra object with sample names and metadata...
achillea_spec
```


## Inspecting and querying spectra

You can check out your spectra object in several ways. For instance, You may want to know how many spectra and how many bands are in there, retrieve the file names, etc. Of course you will need to plot the data, but that topic gets its own section further down.

```{r, eval=TRUE}
# Simply print the object
acer_spectra

# Get the dataset's dimensions
dim(acer_spectra)
```

`spectrolab` also lets you access the individual components of the `spectra`. This is done with the functions `names()` for sample names, `bands()` for band labels, `value()` for the value matrix, and `meta()` for the associated metadata (in case you have any).


```{r, eval=TRUE}
# Vector of all sample names. Note: Duplicated sample names are permitted
n = names(achillea_spec)

# Vector of bands. Note: Duplicated band labels are permitted too (see "Gotchas" below)
w = bands(achillea_spec)

# value matrix
r = value(achillea_spec)

# Metadata. Use simplify = TRUE to get a vector instead of a data.frame
m = meta(achillea_spec, "ssp", simplify = TRUE)
```

## Subsetting spectra

You can subset the `spectra` using a notation *similar* to the `[i, j]` function used in matrices and data.frames. The first argument in `[i, ]` matches *sample names*, whereas the second argument `[ , j]` matches the *band labels* (not column position). Here are some examples of how `[` works in `spectra`:

  - `x[1:3, ]` will keep the first three samples of `x`, i.e. `1:3` are indexes.
  - `x["sp_1", ]` keeps **all** entries in `x` where sample names match `"sp_1"`
  - `x[ , 800:900]` will keep bands labeled `800`, `801`, `802`, ..., `900`.
  - `x[ , bands(x, 800, 900)]` will keep bands between `800` and `900`, including those with non-integer labels, e.g. `876.32`.
  - `x[ , 1:5] ` will **fail**! *bands __cannot__ be subset by column position!*

Subsetting lets you, for instance, exclude noisy regions at the beginning and end of the spectrum or limit the data to specific entries.

```{r, eval = TRUE}
# Subset band regions. Here we know that bands are integers (e.g. 400, 401, ...)
spec_sub_vis = achillea_spec[ , 400:700 ]

# Subset spectra to all entries where sample_name matches "ACHMI_7" or
# get the first three samples
spec_sub_byname = achillea_spec["ACHMI_7", ]
spec_sub_byidx  = achillea_spec[ 1:3, ]
```

The resolution of some spectra may be different from 1nm. In those cases, the best way to subset spectra is using the `min` and `max` arguments for `bands`:

```{r, eval=TRUE}
acer_spectra_trim = acer_spectra[ , bands(acer_spectra, 400, 2400) ]
```

Note that you can (1) subset samples using indexes and (2) use characters or numerics to subset bands. As said before, you cannot use column position to subset bands though.

```{r, eval=TRUE}
# Subsetting samples by indexes works and so does subsetting bands by numerics or characters.
spec_sub_byidx[1, "405"] == spec_sub_byidx[1, 405]
```

*But remember that you CANNOT use a column position to subset bands!*

```{r, eval=TRUE, error=TRUE}
# Something that is obviously a column position, like 2 instead of 401
# (the actual label of the 2nd band in our data), will fail with an
# informative error rather than silently returning the wrong band.
spec_sub_byidx[ , 2]
```

## Plotting

The workhorse function for statically plotting `spectra` is `plot()`. It will jointly plot each spectrum in the `spectra` object. You should be able to pass the usual plot arguments to it, such as `col`, `ylab`, `lwd`, etc.

You can also plot the quantiles of a `spectra` object with `plot_quantile()`. It's second argument, `total_prob`, is the total "mass" that the quantile encompasses. For instance, a `total_prob = 0.95` covers 95% of the variation in the `spectra` object, i.e. it is the `0.025 to 0.975` quantile. The quantile plot can stand alone or be added to a current plot if `add = TRUE`.

The function `plot_regions()` helps shading different spectral regions. `spectrolab` provides a `default_spec_regions()` matrix as an example, but you obviously can customize it for your needs (see the help page for `plot_regions` for details).

```{r, fig.height=2.5, fig.width=7, eval=TRUE}
# Simple spectra plot
oldpar = par(no.readonly = TRUE)

par(mfrow = c(1, 3))
plot(achillea_spec, lwd = 0.75, lty = 1, col = "grey25", main = "All Spectra")

# Stand along quantile plot
plot_quantile(achillea_spec, total_prob = 0.8, col = rgb(1, 0, 0, 0.5), lwd = 0.5, border = TRUE)
title("80% spectral quantile")

# Combined individual spectra, quantiles and shade spectral regions
plot(achillea_spec, lwd = 0.25, lty = 1, col = "grey50", main="Spectra, quantile and regions")
plot_quantile(achillea_spec, total_prob = 0.8, col = rgb(1, 0, 0, 0.25), border = FALSE, add = TRUE)
plot_regions(achillea_spec, regions = default_spec_regions(), add = TRUE)

par(oldpar)
```

Last but not least, spectrolab also allows you to interactively explore spectra through a `shiny` app with the `plot_interactive()` function. If you'd rather work with `ggplot2`, see `autoplot()` in the [Tidy data and ggplot2](tidy_bridge.html) vignette.

## Manipulating sample names, band labels, metadata and value

You may want to edit certain simple attributes of `spectra`, such as making all sample names lowercase. This is easily attainable in `spectrolab`.

```{r, eval=TRUE}
spec_new = achillea_spec

# Replace names with a lowercase version
names(spec_new) = tolower(names(achillea_spec))

# Check the results
names(spec_new)[1:5]
```

If you want to fiddle with the value itself, this is easy, too -- arithmetic
operators (`+`, `-`, `*`, `/`, `^`) and comparisons are all overloaded for
`spectra`.

```{r,  fig.height=3, fig.width=4, fig.align="center", eval=TRUE}
# Scale value by 0.75
spec_new = spec_new * 0.75

# Plot the results
plot(achillea_spec, col = "blue", lwd = 0.75, cex.axis = 0.75)
plot(spec_new, col = "orange", lwd = 0.75, add = TRUE)
```

Or you can also edit or add new metadata to the `spectra` object.
```{r, eval = TRUE}
## Adding metadata to a spectra object: a dummy N content
n_content = rnorm(n = nrow(achillea_spec), mean = 2, sd = 0.5)
meta(achillea_spec, label = "N_percent") = n_content
```

Beyond simple arithmetic, `spectrolab` has verbs for the classic "split many
spectra into groups, summarize, and put back together" workflow:
`aggregate()` collapses spectra into per-group means (or another summary
function); `combine()` row-binds two `spectra` objects (bands must match
exactly, see the [gotcha](#combining-and-aggregating-keep-different-halves-of-your-data) below);
`split()` breaks a `spectra` into a list by a grouping factor; and
`subset_by()` keeps only groups that have between `n_min` and `n_max`
members (handy for dropping species with too few replicates). All four
operate on the whole object -- value, names, and metadata move together.

```{r, eval = TRUE, message = FALSE}
# Average reflectance per species (metadata columns get FUN applied too, when
# possible -- non-numeric metadata, like species names, becomes NA rather
# than erroring, with a message)
spec_species_mean = aggregate(achillea_spec, by = names(achillea_spec), FUN = mean)
dim(spec_species_mean)
```

### Converting a `spectra` object into a matrix or data.frame

It is also possible to convert a `spectra` object to a matrix or data.frame using the `as.matrix()` or `as.data.frame()` functions. This is useful if you want to export your data in a particular format, such as csv.

If you're converting spectra to a matrix, `spectrolab` will (1) place bands in columns, assigning band labels to `colnames`, and (2) samples in rows, assigning sample names to `rownames`. Since `R` imposes strict rules on column name formats and sometimes on row names, `as.matrix()` will try to fix potential dimname issues if `fix_names != "none"`. Note that `as.matrix()` will not keep metadata.

Conversion to data.frame is similar, but keeps the metadata by default (unless you set the `metadata` argument to `FALSE`).

```{r, eval=TRUE}
# Make a matrix from a `spectra` object
spec_as_mat = as.matrix(achillea_spec, fix_names = "none")
spec_as_mat[1:4, 1:3]

# Make a matrix from a `spectra` object
spec_as_df = as.data.frame(achillea_spec, fix_names = "none", metadata = TRUE)
spec_as_df[1:4, 1:5]
```

If you'd rather work with a long/tidy data.frame (one row per
sample-and-band) or a `tibble`, see `to_long()` and `as_tibble()` in the
[Tidy data and ggplot2](tidy_bridge.html) vignette.

## Gotchas

A few behaviors are intentional but easy to be surprised by. Worth reading
once before you hit them in the wild.

### Duplicate sample names *and* duplicate band labels are both allowed

Duplicate sample names have always been allowed (e.g. several scans of the
same individual). As of the current version, **duplicate band labels are
allowed too**, for the same reason: raw SVC files legitimately carry an
identical wavelength on both sides of a sensor overlap (e.g. `975.6` measured
by two detectors) before you've spliced them. Both work the same way when you
subset by the duplicated label: you get *all* matches back, and `spectrolab`
tells you so.

```{r, eval = TRUE}
dup = achillea_spec
bands(dup)[2] = bands(dup)[1]   # force a duplicate band label

# Selecting the duplicated label returns both matching columns -- not an error
dup[1, bands(dup)[1]]
```

Bear in mind that most transform functions (`smooth()`, `normalize()`,
`resample()`, `match_sensors()`, spectral indices, continuum removal, ...)
*require* strictly increasing bands and will error if you hand them raw,
unspliced data with an overlap. Run `match_sensors()` first (see the
[Matching sensor overlap](match_sensors.html) vignette).

### Combining spectra: `spec1 + spec2` vs. `combine()`

`spec1 + spec2` (or `-`, `*`, `/`, `^`) works element-wise once both operands
are `spectra` with identical dimensions and band labels. The result keeps
sample names and metadata **only when both sides agree** -- if they differ,
the result's names/metadata are cleared (with a warning) rather than one side
silently winning:

```{r, eval = TRUE}
s1 = s2 = achillea_spec
meta(s1, "source") = "s1"
meta(s2, "source") = "s2"

r = suppressWarnings(s1 + s2)
ncol(meta(r))   # 0 -- cleared, because s1 and s2's metadata disagreed
```

`combine()`, by contrast, requires bands to match exactly (checked by length
*and* value) and merges the metadata columns from both sides by name, filling
gaps with `NA` -- it is usually what you want when stacking datasets rather
than doing arithmetic on them. `c(s1, s2)` and `rbind(s1, s2)` are shorthands
for `combine()` when you have several objects to stack.

### Comparison operators return a bare matrix, not a `spectra`

`spec1 > spec2`, `spec1 == spec2`, etc. return a logical matrix (samples x
bands) rather than a `spectra` object, since a boolean "spectrum" isn't
meaningful. Only the four arithmetic operators (plus unary `-`/`+`) return
`spectra`.

## Provenance: `quantity()` and `wavelength_unit()`

`read_spectra()` records the physical quantity it read (`"reflectance"` or
`"radiance"`, from the `type` argument) and the wavelength unit (`"nm"` for
every currently supported instrument) as light-touch attributes on the
object:

```{r, eval = TRUE}
quantity(acer_spectra)
wavelength_unit(acer_spectra)
```

Objects without this recorded (e.g. built by hand with `as_spectra()`) return
`NA` for `quantity()` and default to `"nm"` for `wavelength_unit()`. You can
set both yourself with `quantity(x) <- "reflectance"` /
`wavelength_unit(x) <- "nm"`. `combine()` and the arithmetic operators keep
these values when both sides agree and clear them (with a warning) when they
don't -- the same reconciliation rule used for names and metadata above.
Functions that change what the values fundamentally *mean* -- `deriv_spectra()`,
`continuum_removal()` -- clear `quantity` for you, since the output is no
longer raw reflectance/radiance (see the
[Smoothing, derivatives, and continuum removal](preprocessing.html) vignette).

## Where to go next

This vignette covers the basics. Depending on what you're doing next:

* Instrument data with more than one detector? See [Matching sensor overlap](match_sensors.html).
* Need spectra on a different band grid? See [Resampling Spectra](resample.html).
* Want to smooth, differentiate, or remove the continuum from spectra? See [Smoothing, derivatives, and continuum removal](preprocessing.html).
* Computing NDVI, PRI, or a custom two-band index? See [Two-band spectral indices](spectral_indices.html).
* Prefer tidy/long data.frames or `ggplot2`? See [Tidy data and ggplot2](tidy_bridge.html).
