---
title: "Introduction to spectrolab"
author: "Jose Eduardo Meireles, Anna K. Schweiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spectrolab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The goal of `spectrolab` is to provide essential functions to process data from portable spectrometers and a common interface for spectra that other packages can build on. `spectrolab` provides methods to read spectral data files (SVC, ASD and Spectral Evolution), vector normalization, smoothing, matching sensor overlap regions, etc. `spectrolab` introduces the S3 class `spectra`, making it easy to plot and manipulate data and to access additional metadata.

## Installing and loading `spectrolab`

The best way to get `spectrolab` is to install it directly from GitHub. You will need the `devtools` package to do it.

```{r, eval=FALSE}
library("devtools")
install_github("annakat/spectrolab")
```

Assuming that everything went smoothly, you should be able to load `spectrolab` like any other package.

```{r}
library("spectrolab")
```

The source code can be found on our [_GitHub repository_](https://github.com/annakat/spectrolab). Please report any bugs and ask us your questions through the [issue tracker](https://github.com/annakat/spectrolab/issues).

## Reading spectra

There are two ways of getting spectra into R: 1) converting a matrix or data.frame to `spectra` or 2) reading from raw data files (formats: SVC's `sig`, Spectral Evolution's `sed` and ASD's `asd`). Here are a couple examples:

### Create spectra from a matrix

If you already have your spectra in a matrix or data frame (e.g. when you read your data from a .csv file), you can use the function `as.spectra()` to convert it to a `spectra` object. The matrix **must** have samples in rows and wavelengths in columns. The header of the wavelengths columns must be (numeric) wavelength labels. You also should declare which column has the sample names (which are mandatory) using the `name_idx` argument. If other columns are present (other than sample name and reflectances), their indices must be passed to `as.spectra` as the `meta_idxs` argument.

Here is an example using a dataset matrix named `spec_matrix_example` provided by the package.

```{r, eval=TRUE}
# Check out the format of the matrix
spec_matrix_example[1:4, 1:3]

# To convert it to spectra, simply run
spec <- as.spectra(spec_matrix_example, name_idx = 1, meta_idxs = NULL)
```


### Reading spectra: example with SVC's `.sig` files

You shoud use the function `read_spectra()` to read raw spectra files. You can pass a vector of filenames to `read_spectra()`, but it is usually easier to pass the path to the folder where your data are.


```{r, eval=TRUE}

# `dir_path` is the directory where our example datasets live
dir_path <- system.file("extdata", "Acer_example", package = "spectrolab")

# Read .sig files
acer_spectra <- read_spectra(path = dir_path, format = "sig")

# Note that `acer_spectra` is a `spectra` object. 
# You can ensure that this is true using spectrolab's `is_spectra()` function.
is_spectra(acer_spectra)
```



## Inspecting spectra

You can check out your spectra object in several ways. For instance, You may want to know how many spectra and how many bands are in there, retrieve the filenames, and of course you may need to plot the data.

```{r, eval=TRUE}
# Simply print the object
acer_spectra

# Get the dataset's dimensions
dim(acer_spectra)

# Get the sample names
names(acer_spectra)

# and plot the spectra
plot(acer_spectra)
```

When plotting `(acer_spectra)` you will notice that the `Acer_example` data contain one white reference measurement (the horizontal line around 1.0 reflectance) and one obviously bad measurement (with a dip at the NIR shoulder). During the measurement campaign the file names of those two measurements were flagged with "_WR" and "_BAD". You can avoid importing those files to beging with by passing those flags to `read_spectra()`.

```{r, eval=TRUE}

# Check the filenames to see if there are any flags
names(acer_spectra)

# use the `exclude_if_matches` argument to excluded flagged files
acer_spectra <- read_spectra(path = dir_path, format = "sig", exclude_if_matches = c("BAD","WR"))

# and check result
plot(acer_spectra)
```

Flagging "unusual" measurements during data collection is a very good idea. It clearly helps you to quickly sort through your files when it comes to processing. However, you can also exclude bad spectra and outliers later on. We show you how in section `Excluding spectra` futher below. 


### Converting a `spectra` object into a matrix

It is also possible to convert a `spectra` object to a matrix format, using the `as.matrix()` function. This is for example useful if you want to export your data in a particular format, such as .csv. `spectrolab` will (1) place wavelengths in columns, assigning wavelength labels to `colnames`, and (2) samples in rows, assigning sample names to `rownames`. Since `R` imposes strict rules on column name formats and sometimes on row names, `as.matrix()` will try to fix potential dimname issues if `fix_names != "none"`.

```{r, eval=F}
# Make a matrix from a `spectra` object
spec_as_mat = as.matrix(spec, fix_names = "none")
spec_as_mat[1:4, 1:3]
```

## Plotting

The workhorse function for plotting `spectra` is `plot()`. It will jointly plot each spectrum in the `spectra` object. You should be able to pass the usual plot arguments to it, such as `col`, `ylab`, `lwd`, etc.

You can also plot the quantile of a `spectra` object with `plot_quantile()`. It's second argument, `total_prob`, is the total "mass" that the quantile encompasses. For instance, a `total_prob = 0.95` covers 95% of the variation in the `spectra` object, i.e. it is the `0.025 to 0.975` quantile. The quantile plot can stand alone or be added to a current plot if `add = TRUE`.

Last but not least, you can also shade spectral regions with the `plot_regions()` function. `spectrolab` provides a `default_spec_regions()` matrix as an example, but you obviously can customize it for your needs.

```{r, fig.height=2.5, fig.width=8, dev_svg}
# Simple spectra plot
par(mfrow = c(1, 3))
plot(spec, lwd = 0.75, lty = 1, col = "grey25", main = "All Spectra")

# Stand along quantile plot
plot_quantile(spec, total_prob = 0.8, col = rgb(1, 0, 0, 0.5), lwd = 0.5, border = TRUE)
title("80% spectral quantile")

# Combined individual spectra, quantiles and shade spectral regions
plot(spec, lwd = 0.25, lty = 1, col = "grey50", main="Spectra, quantile and regions")
plot_quantile(spec, total_prob = 0.8, col = rgb(1, 0, 0, 0.25), border = FALSE, add = TRUE)
plot_regions(spec, regions = default_spec_regions(), add = TRUE)
```

## Querying

`spectrolab` lets you query the `spectra` object and get summary information. You can easily get sample names with `names()` and wavelength labels with `wavelengths()`. If you need the reflectance, you can retrieve it with the `reflectance()` function.

```{r, eval=F}
# Get the vector of all sample names. Note: Duplicated sample names are permitted
n <- names(spec)
n[1:5]

# Get the vector of wavelengths
w <- wavelengths(spec)
w[1:5]

# or the reflectances in matrix format
r <- reflectance(spec)

# First 10 wavelengths of first 5 species
r[1:5,1:10]

# You can also get the dimensions of your `spectra` object
dim(spec)
```

## Subsetting spectra

You can subset the `spectra` using a notation *similar* to the `[i,j]` function used in matrices and data.frames. The first argument in `[i, ]` matches *sample names*, whereas the second argument `[ ,j]` matches the *wavelength names*. Here are some examples of how `[` works in `spectra`:

  - `x[1:3, ]` will keep the first three samples of `x`, i.e. `1:3` are indexes.
  - `x["sp_1", ]` keeps **all** entries in `x` where sample names match `"sp_1"`
  - `x[ ,800:900]` will keep wavelengths between `800` and `900`.
  - `x[ ,1:5] ` will **fail**!. *wavelengths __cannot__ be subset by indexes!*

Subsetting lets you e.g. exclude noisy regions at the beginning and end of the spectrum or limit the data to specific entries.

```{r, fig.height=2.5, fig.width=6, fig.align="center", eval=F}

# Subset wavelength regions 
spec_sub <-  spec[ ,400:700]

# Check the result
plot(spec_sub)

# Subset spectra to all entries where sample_name matches "species_8"
spec_sp8 <- spec["species_8", ]

# Check the results
dim(spec_sp8)

# Plotting the subset result should work just fine
plot(spec_sp8, col = "red", main = "Species 8 spectra")
plot_quantile(spec_sp8, total_prob = 0.75, add = TRUE,  col = rgb(0.2, 0.2, 0.2, 0.2), border = TRUE)
plot_regions(spec_sp8, default_spec_regions(), col = rgb(1, 0.5, 0, 0.1), add = TRUE)
```

Note that you can (1) subset samples using indexes and (2) use characters or numerics to subset wavelengths. As said before, you cannot use indexes to subset wavelengths though.

```{r, eval=F}
# Subsetting samples by indexes works and so does subsetting wavelengths by numerics or characters.
reflectance(spec_sp8[1,"405"]) == reflectance(spec_sp8[1,405])
```

*But remember that you CANNOT use indexes to subset wavelengths!*

```{r, eval=F}
# Something that is obvioulsy an index, like using 2 instead of 401 (the 2nd band 
# in our case), will fail.
spec_sp8[ ,2]

`Error in i_match_ij_spectra(this = this, i = i, j = j) : Wavelength subscript out of bounds. Use wavelength labels instead of raw indices.`
```

However, `spectrolab` cannot detect if you meant to get the two last bands when
you have 2001 in your dataset and use 2000:2001. It will assume that you 
wanted wavelengths "2000" and "2001".
Bottomline, be very careful not to use indexes to subset wavelengths!


## Manipulating samples and wavelength labels

You may want to edit certain simple attributes of `spectra`, such as make all sample names uppercase. This is easily attainable in `spectrolab`:

```{r, eval=F}
spec_new <- spec

# Replace names with an uppercase version
names(spec_new) <- toupper(names(spec_new))

# Check the results
names(spec_new)[1:5]
```

If you want to fiddle with the reflectance itself, this is easy, too.

```{r,  fig.height=3, fig.width=4, fig.align="center", eval=F}
# Scale reflectance by 0.75
spec_new <- spec_new * 0.75

# Plot the results
plot(spec, col = "blue", lwd = 0.75, cex.axis = 0.75)
plot(spec_new, col = "orange", lwd = 0.75, add = TRUE)
```

## Essential processing 

### Jump correction

Some instruments allow you to preserve the sensor overlap regions. This results in duplicated wavelenghts and a "jump" between the VIS/NIR and NIR/SWIR regions. `spectrolab`'s `jump corrector` removes the overlap regions between the VIS/NIR and NIR/SWIR regions, respectively, and corrects the spectra to allow for better transition between the regions, while keeping the spectral information as close to the original as possible. In principle, the algorithm searches for the wavelengths of closest approach in both overlap regions, cuts the spectrum into three pieces, removes the overlap and finds a multiplier to paste the junks back together. No additional smoothing is applied. As a convention, the NIR region is left unchanged, the VIS and SWIR get adjusted.  

```{r, eval=F}
# Read Acer example spectra 
acer_spectra <- read_spectra(path = dir_path, format = "sig",   exclude_if_matches = c("BAD","WR"))

# and check result
plot(acer_spectra)

# Jump correction 
acer_juco <- jump_corr(acer_spectra)

# This looks better!
plot(acer_juco)
```

### Vector normalisation

Treating each spectrum as a vector and standarizing it to unity length is a great way to remove brightness differences (e.g. due to changing lamp power), while preserving the shape of the spectrum (and thus its information content). Vector normalization is implemented in `spectrolab`'s `normalize()` function. Note: If you want to use *Ms. A's* trait coefficients, you need to use vector normalized spectra, limited to the 400 - 2400 nm wavelength range.

```{r, eval=F}
# Subset jump corrected spectra to 400 - 2400 nm 
acer_sub <- acer_juco[, 400:2400]

# and check result
plot(acer_sub)

# Perform vector normalization
acer_vn <- normalize(acer_sub)

# and check result
plot(acer_vn)
```

### Smoothing at overlap regions ... coming soon

Depending on your analysis, it can be a good idea to smooth remaining noise at the sensor overlap regions. `spectrolab` comes with two smoothing functions: `smoo_visnir()` uses moving average between 970 - 1025 nm. `smoo_nirswir()` uses linear interpolation between 1903 - 1917 nm. Note: *Ms. A* does not do this when predicting traits from spectra. 

```{r, eval=FALSE}
# Smooth only VIS/NIR or NIR/SWIR
acer_smoo1 <- smoo.visnir(acer_juco)
acer_smoo2 <- smoo.nirswir(acer_juco)

# Smooth both regions
acer_smoo <- smoo.nirswir.svc(acer_smoo1)

# Same result
acer_smoo <- smoo.visnir.svc(acer_smoo2)
```

## Excluding spectra 

The `spectrolab` package comes with a couple of options to exclude bad spectra and outliers. The function `excl_shoulder()` focusses on the NIR shoulder (780 nm), a region commonly used to assess the signal to noise ratio in spectral measurements. You can set the upper and lower reflectance limits at 780 nm yourself.

```{r, eval=FALSE}
# Subset jump corrected spectra to 400 - 2400 nm 
acer_sub <- acer_juco[, 400:2400]

# Exclude spectra with reflectances at 780 nm (the "NIR shoulder") <0.3 or >0.65 
acer_sub1 <- excl_shoulder(acer_sub, refl_high = 0.65, refl_low = 0.3)
plot(acer_sub1)
```

Functions `excl_high` and `excl_low` give you more flexibility. Use `excl_high` to exclude measurements where reflectances appear *too high* and `excl_low` to exclude measurements where reflectances appear *too low*. Simply supply the wavelength and reflectance value limit you want to use.

```{r, eval=FALSE}
# First plot your spectra to search for outliers 
plot(acer_sub)

# and add vertical and horizontal guides if you like.
abline(v=1500)
abline(h=0.2)

# Instead of using the NIR shoulder, remove the outlier by defining a lower limit,
# e.g., 0.2 reflectance at 1500 nm, and check the result.
acer_sub2 <- excl_low(acer_sub, refl_low = 0.2, wvl_low = 1500)
plot(acer_sub2)

# Or exclude the same measurement by defining a upper limit, e.g. 0.1 reflectance 
# at 600 nm, and check the result. 
acer_sub3 <- excl_high(acer_sub, refl_high = 0.1, wvl_high = 600)
plot(acer_sub3)
```

Another common feature defining outliers can be dips in the spectrum, e.g. around the NIR shoulder, caused by some sort of sensor failure. To remove such measurements use `excl_dip` and define the maximum allowed difference `refl_diff` between two wavelengths, `wvl_low` which is the shorter wavelength, and `wvl_high` which is the longer wavelength. Use `help(preprocess)` for more information. 
