---
title: "Matching sensor overlap"
author: "Jose Eduardo Meireles"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Matching sensor overlap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library("spectrolab")
knitr::opts_chunk$set(
    echo       = TRUE,
    collapse   = TRUE,
    comment    = "#>",
    fig.align  = "center",
    fig.retina = 2,
    dpi        = 150
)
```

## The problem

Instruments may measure different spectral ranges using different detectors, resulting in abrupt "jumps" in the reflectance or radiance data. In such cases, the regions between sensors need to be matched, i.e. spliced together. Unmatched spectra collected with a 3-sensor instrument, such as the SVC HR1024, may look like this:

```{r, fig.height=8, fig.width=5, fig.align='center', echo=TRUE, out.width="100%"}
# Path to raw (unmatched) spectra
path_raw = system.file("extdata/svc_raw_and_overlap_matched_serbin/SVC_Files/",
                       package = "spectrolab")

# Read spectra as reflectance and radiance
reflect_raw  = read_spectra(path = path_raw, type = "target_reflectance")

radiance_raw = read_spectra(path = path_raw, type = "target_radiance")

# Sensor overlaps marked with vertical dashed lines
lwd = 0.5
cex = 0.7

oldpar = par(no.readonly = TRUE)
par(mfrow = c(2, 1))

plot(reflect_raw, main = "Reflectance",
     lwd = lwd, cex.main = cex, cex.lab = cex, cex.axis = cex)
abline(v = c(990, 1900), col = "red", lty = 2, lwd = lwd)

plot(radiance_raw, main = "Radiance",
     lwd = lwd, cex.main = cex, cex.lab = cex, cex.axis = cex)
abline(v = c(990, 1900), col = "red", lty = 2, lwd = lwd)

par(oldpar)
```

Every instrument -- and every *vendor* -- handles this differently: SVC
**deletes** the overlapping points (optionally after a scalar match), Spectral
Evolution/NaturaSpec **ramp-blends** across the overlap, and ASD ships data
that is *already* joined (it only records where the splice happened, not how).
`spectrolab` therefore offers two ways to splice: a single general-purpose
**legacy algorithm** that works reasonably on any instrument, and a **preset
engine** that mimics each vendor's own approach more closely. Both live behind
`match_sensors()`.

## Where do the splice points come from?

`read_spectra()` captures per-sample provenance about the detector splice at
read time -- into the `sensor_info(x)` attribute -- whenever the source file
records it:

```{r}
sensor_info(reflect_raw)[1:2, ]
```

Here `overlap_mode` is `"Preserve"`, meaning this particular SVC file was
saved *before* splicing, so no splice wavelengths are recorded yet (`splice_1`/
`splice_2` are `NA`) -- you must supply `splice_at` yourself (or estimate it,
see below). Files saved by the instrument *after* on-board splicing (mode
`"Remove"`) *do* record the splice wavelengths, and `match_sensors(x)` will
use them automatically without any arguments -- see
[Provenance-aware defaults](#provenance-aware-defaults).

You can use `plot_interactive()` to zoom into a particular spectral region and decide what the `splice_at` values should be. You can also use the function `guess_splice_at()` to estimate these bands for you, but keep in mind that these are **guesses**.

```{r, fig.height=8, fig.width=5, fig.align='center', echo=TRUE, out.width="100%"}
# Spectrolab's guess of what the splice bands are.
# However, you should also visually inspect the spectra to determine what the
# boundaries between sensors are.
splice_bands_guess = guess_splice_at(reflect_raw)
splice_bands_guess

# Finally, if you know what those sensor bounds should be (say, they're given by
# the manufacturer, or read off the instrument's own header), use those numbers
# instead of spectrolab's guess. This file's own header records 970/1901 (see
# sensor_info() of the matched reference file below) rather than the nominal
# 990/1900 from the manual -- always prefer what the file says over the manual.
splice_bands = c(990, 1900)
```

## The legacy algorithm

Passing `splice_at` (and, optionally, `interpolate_wvl`) without a `method`
runs the original, vendor-agnostic algorithm: it deletes the overlap and
rescales one whole sensor by a single factor computed from a small window on
either side of each splice, linearly ramping that factor to 1 across the
rescaled sensor. It is a reasonable default when you don't know (or don't
need) vendor-specific behavior.

```{r, fig.height=8, fig.width=5, fig.align='center', echo=TRUE, out.width="100%"}
# Match the reflectance and radiance data
reflect_matched = match_sensors(x = reflect_raw, splice_at = splice_bands,
                                interpolate_wvl = c(5, 1))

radiance_matched = match_sensors(x = radiance_raw, splice_at = splice_bands,
                                 interpolate_wvl = c(5, 1))

lwd = 0.5
cex = 0.7

oldpar = par(no.readonly = TRUE)

par(mfrow = c(2, 1))

plot(reflect_raw, main = "Reflectance",
     lwd = lwd, cex.main = cex, cex.lab = cex, cex.axis = cex)

plot(reflect_matched, col = "red", add = TRUE,
     lwd = lwd, cex.main = cex, cex.lab = cex, cex.axis = cex)

plot(radiance_raw, main = "Radiance",
     lwd = lwd, cex.main = cex, cex.lab = cex, cex.axis = cex)

plot(radiance_matched, col = "red", add = TRUE,
     lwd = lwd, cex.main = cex, cex.lab = cex, cex.axis = cex)

par(oldpar)
```

And we can check the results from spectrolab's legacy algorithm against SVC's proprietary matching algorithm (a file the instrument itself already spliced, shipped alongside the raw one for this example).

```{r, fig.height=8, fig.width=5, fig.align='center', echo=TRUE, out.width="100%"}
path_moc = system.file("extdata/svc_raw_and_overlap_matched_serbin/SVC_Files_moc/",
                       package = "spectrolab")

reflect_moc  = read_spectra(path = path_moc, type = "target_reflectance")

radiance_moc = read_spectra(path = path_moc, type = "target_radiance")

lwd = 0.5
cex = 0.7

oldpar = par(no.readonly = TRUE)

par(mfrow = c(2, 1))

plot(reflect_moc, main = "Reflectance", col = "black",
     lwd = lwd, cex.main = cex, cex.lab = cex, cex.axis = cex)

plot(reflect_matched, col = "red", add = TRUE,
     lwd = lwd, cex.main = cex, cex.lab = cex, cex.axis = cex)

plot(radiance_moc, main = "Radiance", col = "black",
     lwd = lwd, cex.main = cex, cex.lab = cex, cex.axis = cex)

plot(radiance_matched, col = "red", add = TRUE,
     lwd = lwd, cex.main = cex, cex.lab = cex, cex.axis = cex)

par(oldpar)
```

## The splice engine: vendor-aware presets

The legacy algorithm above is one whole-sensor scalar match. The splice
*engine* instead decomposes a splice into three independent stages -- exactly
how SVC's own software models it -- so each vendor's behavior can be
expressed as a small, named preset:

```
detect junctions  ->  (optional) GAIN-match one side  ->  JOIN (cut | ramp | concatenate)
```

Pass a preset name via `method`:

| preset        | gain                                   | join        | mimics |
|---------------|----------------------------------------|-------------|--------|
| `"svc"`       | multiplicative, clamped [0.8, 1.2], graded | cut     | SVC HR1024 |
| `"naturaspec"`| SSD-winner (additive or multiplicative) | ramp (linear) | Spectral Evolution / NaturaSpec |
| `"asd"`       | additive, slope-aware (`linear_extrap`) | concatenate | ASD (data already joined) |
| `"cut"` / `"ramp"` / `"concatenate"` | none (generic building blocks) | matching name | -- |

```{r}
reflect_svc_preset = match_sensors(reflect_raw, splice_at = c(970, 1901), method = "svc")

cat("legacy join kept", ncol(reflect_matched), "bands\n")
cat("svc preset kept ", ncol(reflect_svc_preset), "bands (vendor-matched file has", ncol(reflect_moc), ")\n")
```

The `"svc"` preset's clean cut lands much closer to the vendor's own band
count than the legacy ramp does, because it mimics *what SVC actually does*
(delete the overlap) rather than a generic rescale.

For full control -- a non-standard instrument, or tweaking one parameter of a
preset -- build your own `splice_config()` and pass it as `config` (which
overrides `method`):

```{r}
custom_cfg = splice_config(gain_type = "multiplicative", reference = "right",
                           clamp = c(0.8, 1.2), graded = TRUE, join = "cut")
custom_cfg

reflect_custom = match_sensors(reflect_raw, splice_at = c(970, 1901), config = custom_cfg)
```

## Provenance-aware defaults

Because `read_spectra()` records splice provenance, `match_sensors(x)` with no
`splice_at` at all does the sensible thing depending on what it finds:

* **Already spliced** (bands already strictly increasing, e.g. a standard ASD
  spectrum, or an SVC file saved in `"Remove"` mode after on-board splicing):
  returned unchanged, with a message -- it's a safe no-op, not silently
  re-splicing data that doesn't need it.

```{r}
match_sensors(reflect_moc)
```

* **Splice points on file** (an SVC file saved in `"Remove"` mode records
  where it cut): those wavelengths are used automatically.

* **Nothing recorded and nothing supplied** (our raw `"Preserve"`-mode file
  above): a clear error asking you to supply `splice_at`, rather than a guess.

```{r, error=TRUE}
match_sensors(reflect_raw)
```

## Gotchas

* **This is not a bit-exact replay of vendor firmware.** SVC applies its
  matching factor to the reference *and* target radiance with slightly
  different values, so the correction does not fully cancel out of a single
  reflectance column the way it does inside the instrument's own software.
  `spectrolab` reproduces the *removal* structure exactly and applies a
  correct, general gain -- it does not claim to reproduce vendor reflectance
  byte-for-byte. Similarly, the ASD gain is spectrolab's own choice (not a
  documented ASD algorithm), and the Spectral Evolution ramp is a
  single-source (derived) reconstruction. See `?splice_config` for the full
  documentation of these caveats.
* **`interpolate_wvl` only affects the legacy algorithm.** It's silently
  ignored when you pass `method` or `config` -- the engine derives its own
  crossover window from the detected overlap (override it with
  `splice_config(window = c(lo, hi))` if needed).
* **Order of `splice_at` matters less than getting the values right.**
  `match_sensors()` sorts `splice_at` for you, but it cannot tell if you
  handed it the wrong wavelengths for *this* file -- always cross-check
  against `sensor_info(x)` or a visual inspection rather than trusting a
  manual's nominal crossover points (this file's own header says 970/1901,
  not the SVC manual's nominal 990/1900).
* **Everything downstream expects strictly increasing bands.** `smooth()`,
  `normalize()`, `resample()`, continuum removal, and the two-band spectral
  indices all check for this and will error on raw, unspliced data -- run
  `match_sensors()` first.
