---
title: "How to match sensors in raw spectra"
author: "Jose Eduardo Meireles and Anna Schweiger"
output:
  rmarkdown::pdf_document:
vignette: > 
  %\VignetteIndexEntry{Introduction to spectrolab}
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library("spectrolab")
knitr::opts_chunk$set(echo = TRUE)
```

Raw spectra may have abrupt reflectance 'jumps' in the interface between sensors, and those need to be matched (spliced). In a raw spectrum collected with a 3-sensor intrument (say SVC), these regions may look like this:

```{r, fig.height=3.5, fig.width=4, fig.align='center'}
plot(spec_with_jump, cex.lab = 0.5, cex.axis = 0.6)

sensors  = matrix(data     = c(400, 970, 971, 1909, 1910, 2500), 
                  ncol     = 3,
                  dimnames = list(c("begin", "end"), c("S1", "S2", "S3"))) 
plot_regions(spec_with_jump, regions = sensors, border = F, lwd = 0.15, lty = 3,
             col = adjustcolor(c("green", "blue", "orange"), 0.035), cex_label = 0.75)
```

In this case, sensor 1 goes from 400 to 970nm, sensor 2 from 971 to 1909 and sensor 3 from 1910 to 2500. These boundaries will vary depending on the individual instument, so you'll have to plot and zoom your spectra to figure out where the sensor limits are.

Aside from a spectra object, the `match_sensors` function takes a few other important agruments: 

1. `splice_at`: A numberic vector with the start wavelengths of the 2nd and 3rd sensors. As said before, you'll need to plot your spectra to dins our what these wavelenegths are. This argument must have length 1 (2 sensors) or length 2 (3 sensors).

2. `fixed_sensor`: Which sensor should be used as a reference. If `splice_at` is length 2, `fixed_sensor` must be the 2nd one. Otherwise, you can choose to fix either the 1st or 2nd sensor. 

3. `interpolate_wvl`: Extent around `splice_at` wavelengths over which the matching algorithm computes the conversion factors.

Here are a few combinations of different matching parameters to give you an idea of how the function is called and what the ourtputs look like:


```{r, warning=FALSE}
a = match_sensors(spec_with_jump, splice_at = c(971, 1910), fixed_sensor = 2,
                  interpolate_wvl = c(2, 2))
b = match_sensors(spec_with_jump, splice_at = c(971, 1910), fixed_sensor = 1,
                  interpolate_wvl = c(2, 2))
c = match_sensors(spec_with_jump, splice_at = c(971),       fixed_sensor = 2,
                  interpolate_wvl = c(2, 2))
d = match_sensors(spec_with_jump, splice_at = c(971),       fixed_sensor = 1,
                  interpolate_wvl = c(2, 2))
e = match_sensors(spec_with_jump, splice_at = c(1910),      fixed_sensor = 2,
                  interpolate_wvl = c(2, 2))
f = match_sensors(spec_with_jump, splice_at = c(1910),      fixed_sensor = 1,
                  interpolate_wvl = c(2, 2))
```


```{r, fig.height=6, fig.width=4, fig.align='center', echo=FALSE}

par(mfrow = c(3, 2))

plot(spec_with_jump, main = "splice_at = c(971, 1910)\nfixed_sensor = 2 ", cex.main = 1, col = "grey35")
plot(a, add = TRUE, col = "red")
plot_regions(spec_with_jump, regions = sensors, border = F, lwd = 0.15, lty = 3,
             col = adjustcolor(c("green", "blue", "orange"), 0.035), cex_label = 0.65)

plot(spec_with_jump, main = "splice_at = c(971, 1910)\nfixed_sensor = 1 ", cex.main = 1, col = "grey35")
plot(b, add = TRUE, col = "red")
plot_regions(spec_with_jump, regions = sensors, border = F, lwd = 0.15, lty = 3,
             col = adjustcolor(c("green", "blue", "orange"), 0.035), cex_label = 0.65)


plot(spec_with_jump, main = "splice_at = 971\nfixed_sensor = 2 ", cex.main = 1, col = "grey35")
plot(c, add = TRUE, col = "red")
plot_regions(spec_with_jump, regions = sensors, border = F, lwd = 0.15, lty = 3,
             col = adjustcolor(c("green", "blue", "orange"), 0.035), cex_label = 0.65)


plot(spec_with_jump, main = "splice_at = 971\nfixed_sensor = 1 ", cex.main = 1, col = "grey35")
plot(d, add = TRUE, col = "red")
plot_regions(spec_with_jump, regions = sensors, border = F, lwd = 0.15, lty = 3,
             col = adjustcolor(c("green", "blue", "orange"), 0.035), cex_label = 0.65)

plot(spec_with_jump, main = "splice_at = 1910\nfixed_sensor = 2 ", cex.main = 1, col = "grey35",
     ylim = c(0, 0.65))
plot(e, add = TRUE, col = "red")
plot_regions(spec_with_jump, regions = sensors, border = F, lwd = 0.15, lty = 3,
             col = adjustcolor(c("green", "blue", "orange"), 0.035), cex_label = 0.65)


plot(spec_with_jump, main = "splice_at = 1910\nfixed_sensor = 1 ", cex.main = 1, col = "grey35",
     ylim = c(0, 0.65))
plot(f, add = TRUE, col = "red")
plot_regions(spec_with_jump, regions = sensors, border = F, lwd = 0.15, lty = 3,
             col = adjustcolor(c("green", "blue", "orange"), 0.035), cex_label = 0.65)

```

