---
title: "Smoothing, derivatives, and continuum removal"
author: "Jose Eduardo Meireles"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Smoothing, derivatives, and continuum removal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse   = TRUE,
    comment    = "#>",
    fig.align  = "center",
    fig.retina = 2,
    dpi        = 150
)
```

This vignette covers `spectrolab`'s spectral preprocessing tools: smoothing
(including a Savitzky-Golay filter), spectral derivatives, and continuum
removal. All of them are transform functions -- they take a `spectra` object
and return one -- and all of them **require strictly increasing bands**
(splice sensor overlaps with `match_sensors()` first if needed; see the
[Matching sensor overlap](match_sensors.html) vignette).

```{r, message = FALSE}
library(spectrolab)

spec = as_spectra(spec_matrix_example, name_idx = 1)
one  = spec[1, ]   # a single spectrum, used for the "before/after" plots below
```

## Smoothing

`smooth()` dispatches to four methods via its `method` argument:

| `method`          | function             | idea |
|-------------------|----------------------|------|
| `"gaussian"` (default) | `smooth_fwhm()` | resample onto the same bands using a wider Gaussian FWHM (see the [Resampling Spectra](resample.html) vignette) |
| `"spline"`         | `smooth_spline()`    | fit `stats::smooth.spline()` per sample |
| `"moving_average"` | `smooth_moving_avg()`| simple moving average over `n` adjacent bands |
| `"sgolay"`         | `smooth_sgolay()`    | Savitzky-Golay filter: a local polynomial fit over a moving window (requires the `signal` package) |

```{r, fig.height=3.5, fig.width=7}
s_gauss = smooth(one, method = "gaussian")
s_sg    = suppressMessages(smooth(one, method = "sgolay"))

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(one, lwd = 1, col = "grey60", main = "Gaussian (default)")
plot(s_gauss, add = TRUE, col = "firebrick", lwd = 1)

plot(one, lwd = 1, col = "grey60", main = "Savitzky-Golay")
plot(s_sg, add = TRUE, col = "steelblue", lwd = 1)

par(oldpar)
```

### The Savitzky-Golay filter in more detail

`smooth_sgolay(x, p, n)` fits a degree-`p` polynomial over a moving window of
`n` bands and replaces the center value with the fitted one. `n` must be odd
and greater than `p`; the default, `n = p + 3 - p %% 2`, is deliberately
small (5 bands, for the default `p = 3`) so it does very little smoothing
until you widen the window:

```{r, fig.height=3.5, fig.width=7}
s_narrow = smooth_sgolay(one, p = 3, n = 5)    # default-sized window
s_wide   = smooth_sgolay(one, p = 3, n = 51)   # a much wider window

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(one, lwd = 1, col = "grey60", main = "n = 5 (default-sized)")
plot(s_narrow, add = TRUE, col = "steelblue", lwd = 1)

plot(one, lwd = 1, col = "grey60", main = "n = 51")
plot(s_wide, add = TRUE, col = "steelblue", lwd = 1)

par(oldpar)
```

Unlike a simple moving average, the Savitzky-Golay filter has full-length
output at the edges (no bands are dropped there) because its polynomial fit
is defined at every point in the window, including the boundary -- see the
[Gotchas](#gotchas) below for how `smooth_moving_avg()` differs.

## Spectral derivatives

`deriv_spectra(x, order)` computes the order-`order` derivative of each
spectrum using the same Savitzky-Golay machinery (also requires `signal`).
Derivatives are a classic way to emphasize subtle inflection points -- like
the position of the red edge -- that are hard to see in raw reflectance.

```{r, fig.height=3.5, fig.width=7}
d1 = suppressMessages(deriv_spectra(one, order = 1))
d2 = suppressMessages(deriv_spectra(one, order = 2))

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(d1, main = "1st derivative", col = "darkgreen", ylab = "dReflectance/dWavelength")
abline(h = 0, col = "grey", lty = 2)

plot(d2, main = "2nd derivative", col = "purple", ylab = "d2Reflectance/dWavelength2")
abline(h = 0, col = "grey", lty = 2)

par(oldpar)
```

`deriv_spectra()` (like `smooth_sgolay()`) emits a message reminding you that
the output is **no longer reflectance/radiance** -- the y-axis units have
changed. Pass `quiet = TRUE` once you know that.

## Continuum removal

The "continuum" of a spectrum is its upper convex hull in wavelength/value
space -- essentially the smooth envelope you'd draw connecting the local
reflectance peaks. Dividing a spectrum by its continuum, `continuum_removal()`,
suppresses broad absorption features and emphasizes the depth of local ones
(a standard technique in mineral and vegetation spectroscopy for comparing
absorption features across samples that otherwise differ in overall
brightness/albedo).

```{r, fig.height=3.5, fig.width=7}
one_cr = continuum_removal(one)

# Compute the hull the same way continuum_removal() does internally, purely
# to visualize it -- this is NOT required to use continuum_removal() itself.
w = bands(one)
r = as.numeric(value(one))
h = grDevices::chull(w, r)
chord = r[[1]] + (r[[length(r)]] - r[[1]]) * (w[h] - w[[1]]) / (w[[length(w)]] - w[[1]])
h = sort(h[r[h] >= chord - sqrt(.Machine$double.eps)])
hull = stats::approx(w[h], r[h], xout = w)$y

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(one, lwd = 1, col = "grey30", main = "Spectrum and its continuum")
lines(w, hull, col = "firebrick", lwd = 1.5)

plot(one_cr, lwd = 1, col = "darkorange", main = "Continuum-removed")
abline(h = 1, col = "grey", lty = 2)

par(oldpar)
```

Continuum-removed values are always at or below 1 by construction (they are
a spectrum divided by an envelope that sits at or above it everywhere), and
they are, like derivatives, no longer reflectance.

## Gotchas

* **The `signal` package is required for `smooth_sgolay()` and
  `deriv_spectra()`**, but it's only in `Suggests` -- if it isn't installed
  you'll get a clear error telling you to install it, rather than a cryptic
  one.
* **The method name is `"sgolay"`**, not `"savitzky_golay"` or `"savgol"` --
  matching the naming used by the underlying `signal::sgolay()`/`sgolayfilt()`
  functions it wraps.
* **`smooth_moving_avg()` can silently drop bands.** Its moving average
  produces `NA` at the edges (there aren't enough neighbors to average
  there), and those edge bands are then *removed* from the result -- your
  output can have fewer bands than the input. The dropped values are stashed
  into metadata (`save_bands_to_meta = TRUE`, the default) so they aren't
  lost, just relocated. `smooth_sgolay()` does not have this issue -- its
  polynomial fit is defined at the edges too.
* **Derivative and continuum-removed values are not reflectance/radiance
  anymore**, even though they're still stored in a `spectra` object and will
  happily flow into any function that doesn't check units (e.g.
  `normalize()`, the two-band spectral indices). `spectrolab` does not track
  this for you -- keep your own naming/metadata discipline once you start
  transforming values this way.
* **Choose the Savitzky-Golay window (`n`) relative to your spectral
  resolution**, not in the abstract: 5 bands means something very different
  on 1nm-resolution data than on 10nm data. The default is intentionally
  small; widen it deliberately, and check the result visually as we did
  above.
