---
title: "Smoothing and derivatives"
author: "Jose Eduardo Meireles"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Smoothing and derivatives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse   = TRUE,
    comment    = "#>",
    fig.align  = "center",
    fig.retina = 2,
    dpi        = 150
)
```

This vignette covers `spectrolab`'s spectral preprocessing tools: smoothing
(including a Savitzky-Golay filter) and spectral derivatives. All of them are transform functions -- they take a `spectra` object
and return one -- and all of them **require strictly increasing bands**
(splice sensor overlaps with `match_sensors()` first if needed; see the
[Matching sensor overlap](match_sensors.html) vignette).

```{r, message = FALSE}
library(spectrolab)

spec = as_spectra(spec_matrix_example, name_idx = 1)
one  = spec[1, ]   # a single spectrum, used for the "before/after" plots below
```

## Smoothing

`smooth()` dispatches to four methods via its `method` argument:

| `method`          | function             | idea |
|-------------------|----------------------|------|
| `"gaussian"` (default) | `smooth_fwhm()` | resample onto the same bands using a wider Gaussian FWHM (see the [Resampling Spectra](resample.html) vignette) |
| `"spline"`         | `smooth_spline()`    | fit `stats::smooth.spline()` per sample |
| `"moving_average"` | `smooth_moving_avg()`| simple moving average over `n` adjacent bands |
| `"sgolay"`         | `smooth_sgolay()`    | Savitzky-Golay filter: a local polynomial fit over a moving window (requires the `signal` package) |

```{r, fig.height=3.5, fig.width=7}
s_gauss = smooth(one, method = "gaussian")
s_sg    = suppressMessages(smooth(one, method = "sgolay"))

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(one, lwd = 1, col = "grey60", main = "Gaussian (default)")
plot(s_gauss, add = TRUE, col = "firebrick", lwd = 1)

plot(one, lwd = 1, col = "grey60", main = "Savitzky-Golay")
plot(s_sg, add = TRUE, col = "steelblue", lwd = 1)

par(oldpar)
```

### The Savitzky-Golay filter in more detail

`smooth_sgolay(x, p, n)` fits a degree-`p` polynomial over a moving window of
`n` bands and replaces the center value with the fitted one. `n` must be odd
and greater than `p`; the default, `n = p + 3 - p %% 2`, is deliberately
small (5 bands, for the default `p = 3`) so it does very little smoothing
until you widen the window:

```{r, fig.height=3.5, fig.width=7}
s_narrow = smooth_sgolay(one, p = 3, n = 5)    # default-sized window
s_wide   = smooth_sgolay(one, p = 3, n = 51)   # a much wider window

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(one, lwd = 1, col = "grey60", main = "n = 5 (default-sized)")
plot(s_narrow, add = TRUE, col = "steelblue", lwd = 1)

plot(one, lwd = 1, col = "grey60", main = "n = 51")
plot(s_wide, add = TRUE, col = "steelblue", lwd = 1)

par(oldpar)
```

Unlike a simple moving average, the Savitzky-Golay filter has full-length
output at the edges (no bands are dropped there) because its polynomial fit
is defined at every point in the window, including the boundary -- see the
[Gotchas](#gotchas) below for how `smooth_moving_avg()` differs.

## Spectral derivatives

`deriv_spectra(x, order)` computes the order-`order` derivative of each
spectrum using the same Savitzky-Golay machinery (also requires `signal`).
Derivatives are a classic way to emphasize subtle inflection points -- like
the position of the red edge -- that are hard to see in raw reflectance.

```{r, fig.height=3.5, fig.width=7}
d1 = suppressMessages(deriv_spectra(one, order = 1))
d2 = suppressMessages(deriv_spectra(one, order = 2))

oldpar = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

plot(d1, main = "1st derivative", col = "darkgreen", ylab = "dReflectance/dWavelength")
abline(h = 0, col = "grey", lty = 2)

plot(d2, main = "2nd derivative", col = "purple", ylab = "d2Reflectance/dWavelength2")
abline(h = 0, col = "grey", lty = 2)

par(oldpar)
```

`deriv_spectra()` (like `smooth_sgolay()`) emits a message reminding you that
the output is **no longer reflectance/radiance** -- the y-axis units have
changed. Pass `quiet = TRUE` once you know that.

## Gotchas

* **The `signal` package is required for `smooth_sgolay()` and
  `deriv_spectra()`**, but it's only in `Suggests` -- if it isn't installed
  you'll get a clear error telling you to install it, rather than a cryptic
  one.
* **The method name is `"sgolay"`**, not `"savitzky_golay"` or `"savgol"` --
  matching the naming used by the underlying `signal::sgolay()`/`sgolayfilt()`
  functions it wraps.
* **`smooth_moving_avg()` can silently drop bands.** Its moving average
  produces `NA` at the edges (there aren't enough neighbors to average
  there), and those edge bands are then *removed* from the result -- your
  output can have fewer bands than the input. The dropped values are stashed
  into metadata (`save_bands_to_meta = TRUE`, the default) so they aren't
  lost, just relocated. `smooth_sgolay()` does not have this issue -- its
  polynomial fit is defined at the edges too.
* **Derivative values are not reflectance/radiance anymore.**
  `deriv_spectra()` (and `smooth_sgolay()` with a derivative order > 0) change
  what the y-axis means. The values still flow into any downstream function
  (e.g. `normalize()`, the two-band spectral indices) -- nothing blocks further
  processing, so it is on you to remember that the units have changed.
* **Choose the Savitzky-Golay window (`n`) relative to your spectral
  resolution**, not in the abstract: 5 bands means something very different
  on 1nm-resolution data than on 10nm data. The default is intentionally
  small; widen it deliberately, and check the result visually as we did
  above.
