---
title: "Resampling Spectra"
author: "Jose Eduardo Meireles"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Resampling Spectra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse   = TRUE,
    comment    = "#>",
    fig.align  = "center",
    fig.retina = 2,
    dpi        = 150
)

library(spectrolab)
```

## What resampling does

Different instruments (or the same instrument at different settings) report
spectra on different band grids -- different starting wavelength, spacing, or
resampling to your reference dataset's own grid before comparing/joining. But
resampling isn't just linear interpolation: `resample()` models each new band
as a **Gaussian-weighted average of nearby original bands**, using the **Full
Width at Half Maximum (FWHM)** of each new band as that Gaussian's spread.
This mimics how a real spectrometer's detector integrates light across a
small band, rather than pretending the instrument measured an infinitely
narrow slice at each wavelength. Bands must be strictly increasing already
(`match_sensors()` first, if needed -- see the
[Matching sensor overlap](match_sensors.html) vignette).

## A quick example

```{r, eval=TRUE, out.width="100%"}
dir_path = system.file("extdata/svc_raw_and_overlap_matched_serbin/SVC_Files_moc", package = "spectrolab")

s_raw     = read_spectra(dir_path)

new_bands = seq(400, 2400, 5)
fwhm      = make_fwhm(s_raw, new_bands)

s_resamp  = resample(s_raw, new_bands = new_bands, fwhm = fwhm)

plot(s_raw, lwd = 0.5)
plot(s_resamp, add = TRUE, lwd = 0.25, col = "red")
```

## Where the FWHM comes from

`resample()` needs one FWHM value per new band -- `make_fwhm()` estimates a
sensible one for you from the spacing of the *original* bands, resampled (via
the same Gaussian model) onto the *new* band grid, so a new band is never
narrower than the data that feeds it.

```{r}
# FWHM computed from make_fwhm() is one value per new band
length(fwhm) == length(new_bands)
range(fwhm)
```

You can supply your own FWHM instead (e.g. straight from the instrument's
specs) -- a single number recycled across all new bands, or one value per new
band.

## Gotchas

### `make_fwhm()`'s default `k` bins the FWHM into just 3 clusters

By default `make_fwhm(spec, new_bands, k = 3)` runs a k-means clustering step
over the estimated FWHM values and replaces each one with its cluster's
center. This keeps the *number of distinct resampling kernels* small (cheaper
to compute) when your new bands have only a couple of natural resolutions
(e.g. "5nm in the VIS/NIR, 10nm in the SWIR"), but it means the FWHM you get
back is **not** a smooth, continuous function of wavelength even when the true
one would be -- and it is easy to mistake the clustering artifact for a
property of your data. The effect is invisible when your new bands are evenly
spaced (there's only one true resolution to begin with) but obvious when they
are not:

```{r}
# Dense in the VIS/NIR (2nm), sparse in the SWIR (10nm)
uneven_bands = c(seq(400, 1000, 2), seq(1005, 2400, 10))

fwhm_default = make_fwhm(s_raw, uneven_bands)          # k = 3 (default)
fwhm_full    = make_fwhm(s_raw, uneven_bands, k = 0)   # no clustering

length(unique(round(fwhm_default, 3)))  # only 3 distinct values...
length(unique(round(fwhm_full, 3)))     # ...vs the true band-to-band variation
```

Set `k = 0` if you want the uncollapsed, per-band FWHM estimate instead.

### Requested bands outside the original range are silently trimmed

`resample()` will not invent data beyond the edges of what you measured. Any
`new_bands` that fall (approximately) outside the original band range are
dropped, with a message -- they are not extrapolated, and the function does
not error.

```{r}
# Ask for bands starting a bit before and ending a bit after the original range
too_wide = seq(330, 2525, 5)
s_trim   = resample(s_raw, new_bands = too_wide, fwhm = make_fwhm(s_raw, too_wide))

range(bands(s_raw))     # what we actually measured
range(bands(s_trim))    # what we got back -- trimmed to fit
```

(Asking for bands *far* outside the original range -- tens of nanometers or
more -- doesn't just get trimmed: the Gaussian weights underlying
`make_fwhm()` underflow to zero and it errors out rather than returning
nonsense. Either way, `resample()` will not extrapolate for you.)

### Upsampling doesn't add real spectral detail

Asking for a finer band spacing than your instrument actually has (e.g.
resampling 1nm data to 0.1nm) will run without error and produce smoothly
interpolated values, but it cannot recover detail the original sensor never
recorded. Treat upsampled data as a smoother *display* of your original
measurement, not as new information -- avoid using upsampled resolution to
justify statistical claims that depend on it.
