---
title: "Two-band spectral indices"
author: "Jose Eduardo Meireles"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Two-band spectral indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse   = TRUE,
    comment    = "#>",
    fig.align  = "center",
    fig.retina = 2,
    dpi        = 150
)
```

A large family of vegetation and mineral indices are just a **normalized
difference between two bands**: `(R[band1] - R[band2]) / (R[band1] +
R[band2])`. NDVI and PRI are two well-known examples, but the formula itself
doesn't care which bands you plug in. Rather than exporting a growing list of
one-off named functions, `spectrolab` exports a single *factory*,
`make_spectral_index()`, and ships two ready-made indices built with it,
reachable through the `spectral_index` list.

```{r, message = FALSE}
library(spectrolab)
spec = as_spectra(spec_matrix_example, name_idx = 1)
```

## The built-in indices

```{r}
names(spectral_index)

# Call an entry directly on a spectra object
head(spectral_index$ndvi(spec))
head(spectral_index$pri(spec))
```

* `spectral_index$ndvi` -- Normalized Difference Vegetation Index, `(R800 - R680) / (R800 + R680)`
* `spectral_index$pri`  -- Photochemical Reflectance Index, `(R570 - R531) / (R570 + R531)`

Both return a named numeric vector, one value per sample (names come from
`names(x)`), not a `spectra` object -- an index is a per-sample summary
statistic, not a spectrum.

To compute several at once, call the ones you want and assemble a data.frame
yourself -- there is no wrapper for this because it is a one-liner:

```{r}
idx = data.frame(ndvi = spectral_index$ndvi(spec),
                 pri  = spectral_index$pri(spec))
head(idx)
```

```{r, fig.height=4, fig.width=4.5}
plot(idx$ndvi, idx$pri, xlab = "NDVI", ylab = "PRI", pch = 16,
     col = as.factor(names(spec)))
```

## Building your own index

`make_spectral_index(band1, band2, tolerance = 1)` returns a function of
`(x, tolerance)` -- exactly how `spectral_index$ndvi` and `spectral_index$pri`
themselves are defined. Use it directly for a custom two-band index, without
adding a new name to the package's namespace:

```{r}
# A water-band index, e.g. R970 vs a reference band at R900
wbi = make_spectral_index(970, 900)
head(wbi(spec))
```

If you use a custom index often, keep the function around (or add it to your
own named list, mirroring `spectral_index`) rather than recomputing
`make_spectral_index()` on every call -- it's cheap either way, but a
descriptive variable name documents intent.

## Gotchas

* **Bands must be strictly increasing.** Like the other transform functions,
  indices error on raw, unspliced data -- run `match_sensors()` first (see
  the [Matching sensor overlap](match_sensors.html) vignette).
* **"Nearest band within tolerance", not "exact band".** Your data's bands
  rarely land exactly on 800.000 or 680.000; `spectral_index` functions find
  the *nearest available band* to the requested wavelength and use it,
  erroring only if nothing is within `tolerance` (default 1, in the same
  units as your bands -- usually nm). On coarse-resolution data this means
  the band actually used could be a few tenths of a nm away from what you
  asked for; widen `tolerance` deliberately if your data's resolution
  requires it, rather than being surprised by an error.

```{r, error = TRUE}
# spec is at 1nm resolution. Asking for an off-grid band (970.5) with
# tolerance = 0 fails rather than silently snapping to the nearest one...
off_grid = make_spectral_index(970.5, 900)
off_grid(spec, tolerance = 0)

# ...while the default tolerance (1) is enough to snap to band 970 and succeed
head(off_grid(spec))
```

* **The band wavelengths are captured when you call `make_spectral_index()`,
  not re-evaluated later.** `wbi = make_spectral_index(970, 900)` always
  compares bands 970 and 900, even if you later change what `970` or `900`
  mean in your analysis -- there's no lazy/symbolic re-binding. Call
  `make_spectral_index()` again if you need different bands.
* **An index is a summary, not a spectrum.** The output has one value per
  *sample*, not one value per *band* -- don't expect `plot()` (which expects
  a `spectra` object) to work on it.
