---
title: "Tidy data and ggplot2"
author: "Jose Eduardo Meireles"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Tidy data and ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse   = TRUE,
    comment    = "#>",
    fig.align  = "center",
    fig.retina = 2,
    dpi        = 150
)
```

`spectrolab`'s native representation -- one row per sample, one column per
band (see `as.matrix()`/`as.data.frame()` in the
[Introduction](introduction_to_spectrolab.html) vignette) -- is *wide*. That's
the right shape for the linear-algebra-heavy operations the rest of the
package does (resampling, splicing, smoothing), but it's the wrong shape for
`ggplot2` and most tidyverse workflows, which expect *long* data: one row per
observation, here one row per sample-and-band. This vignette covers the
bridge between the two.

```{r, message = FALSE}
library(spectrolab)
spec = as_spectra(spec_matrix_example, name_idx = 1)
```

## `to_long()`: dependency-free reshaping

`to_long()` uses only base R -- no tidyverse packages required -- and returns
a long data.frame with `sample_name`, `band`, `value`, and (by default) your
metadata columns repeated once per band.

```{r}
long = to_long(spec)
dim(long)     # one row per sample x band: 50 samples * 2101 bands
head(long)
```

Metadata is included by default; drop it with `metadata = FALSE` if you don't
need it (smaller output, and metadata columns to the same value get repeated
once per band).

```{r}
long_no_meta = to_long(spec, metadata = FALSE)
names(long_no_meta)
```

## `as_tibble()`: the same thing, as a tibble

If you have the `tibble` package installed, `as_tibble()` is defined for
`spectra` objects and wraps `to_long()`:

```{r, eval = requireNamespace("tibble", quietly = TRUE)}
tbl = tibble::as_tibble(spec)
tbl
```

## `autoplot()`: a `ggplot2` view

If you have `ggplot2` installed, `autoplot()` draws one line per sample on
the wavelength/value plane -- a `ggplot2` equivalent of `plot.spectra()` (see
the [Introduction](introduction_to_spectrolab.html) vignette), useful when you
want to layer `ggplot2` facets, themes, or additional geoms on top.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.height = 3.5, fig.width = 6}
ggplot2::autoplot(spec) + ggplot2::theme_minimal()
```

Because `autoplot()` returns a plain `ggplot` object, it composes with the
rest of the grammar normally -- facet by a metadata column, recolor by group,
add a theme, etc.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.height = 3.5, fig.width = 6}
long = to_long(spec)

ggplot2::ggplot(long, ggplot2::aes(x = band, y = value, group = sample_name, color = sample_name)) +
    ggplot2::geom_line(show.legend = FALSE, alpha = 0.5) +
    ggplot2::theme_minimal() +
    ggplot2::labs(x = "wavelength (nm)", y = "reflectance")
```

## Gotchas

* **`tibble` and `ggplot2` are optional.** Both are in `Suggests`, not
  `Imports` -- `to_long()` never needs them, but `as_tibble()` and
  `autoplot()` will stop with a clear message telling you which package to
  install if it's missing, rather than a cryptic "could not find function"
  error.
* **Long format is much bigger than wide.** Reshaping to long duplicates
  `sample_name` (and every metadata column) once per band -- for this
  vignette's small example (50 samples x 2101 bands) that's 105,050 rows for
  three columns. On a real dataset with many samples and full-resolution
  (2000+ band) spectra, `to_long()`/`as_tibble()` can produce a data.frame far
  larger than the original `spectra` object; use `metadata = FALSE` if you
  don't need the repeated metadata columns, and prefer working in wide format
  (`value()`, `as.matrix()`) for anything performance-sensitive.
* **This bridge is one-way.** There is no `as_spectra.tbl_df`/`from_long()`
  that reconstructs a `spectra` object from long data -- go back through
  `as_spectra()` on a wide matrix/data.frame (e.g. after a `tidyr::pivot_wider()`)
  if you need to return to `spectra`.
