---
title: "Visualizing spectral variation"
author: "Jose Eduardo Meireles"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Visualizing spectral variation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
source("_common.R")
library("spectrolab")
```

A single reflectance curve is easy to read; a few hundred of them stacked on
one plot are not. This vignette is about *seeing* variation: coloring spectra
by a grouping, summarizing spread with quantile bands, exposing the strong
band-to-band redundancy that makes spectra behave like a handful of underlying
signals, and extracting the dominant axes of variation with PCA. It uses both
the base-R plotting in `spectrolab` and the tidy/`ggplot2` bridge
(`to_long()`, `as_tibble()`, `autoplot()`).

Throughout we use the built-in `spec_matrix_example`, a **simulated** dataset
of 50 spectra whose sample *names* encode 10 "species" -- a convenient stand-in
for a categorical grouping.

```{r}
s = as_spectra(spec_matrix_example, name_idx = 1)
dim(s)                     # 50 samples x 2101 bands (400-2500 nm)
length(unique(names(s)))   # 10 groups, carried in the sample names
```

The `_common.R` helper (sourced above) supplies the shared visual language used
in every vignette: a colorblind-safe categorical palette (`spec_pal`),
sequential/diverging ramps (`spec_seq()`, `spec_div()`), light region tints
(`spec_region_tints`), a `ggplot2` theme (`theme_spectra()`), and two grouping
helpers. `species_cols(x)` returns a per-sample color vector; `group_palette(x)`
returns the same mapping plus the named palette and levels needed to build a
legend. Both cap the grouping at 8 colors and fold the rest into `"Other"`.

## Coloring spectra by a grouping (base R)

The workhorse is `plot.spectra()`. Passing a per-sample color vector from
`species_cols()` turns an undifferentiated tangle of lines into a
group-structured one. We shade the canonical spectral regions underneath with
`plot_regions()` and build a legend from the named palette in
`group_palette()`.

```{r, fig.width = 7, fig.height = 4.5}
gp = group_palette(s)

plot(s, col = species_cols(s), lwd = 1, ylab = "reflectance",
     xlab = "wavelength (nm)")
plot_regions(s, col = spec_region_tints)
legend("topleft", legend = names(gp$pal), col = gp$pal,
       lwd = 2, bty = "n", cex = 0.8, ncol = 2)
```

Even colored, 50 overlapping curves hide the *within-group* spread. A cleaner
summary draws each group as a shaded quantile band instead of individual lines.
We map every sample back to its (folded) group, then loop, calling
`plot_quantile(add = TRUE)` on each subset with a translucent fill of that
group's color.

```{r, fig.width = 7, fig.height = 4.5}
grp_of = stats::setNames(names(gp$pal)[match(gp$cols, gp$pal)], names(s))

plot(s, type = "n", ylab = "reflectance", xlab = "wavelength (nm)")
plot_regions(s, col = spec_region_tints)

for(g in names(gp$pal)){
    idx = which(grp_of == g)
    if(length(idx) >= 2){
        plot_quantile(s[idx, ], total_prob = 0.9, add = TRUE, border = FALSE,
                      col = grDevices::adjustcolor(gp$pal[[g]], 0.25))
    }
}
legend("topleft", legend = names(gp$pal), fill = gp$pal, bty = "n",
       cex = 0.8, ncol = 2, border = NA)
```

Each band spans the central 90% of its group's reflectance at every wavelength.
Overlapping bands mean the groups are not cleanly separable there; narrow,
non-overlapping bands (e.g. the red edge near 700 nm) mark the wavelengths that
best discriminate the groups.

## The tidy path: ggplot2

For grammar-of-graphics workflows, reshape to long format with `to_long()`
(base R, no tidyverse required) and hand the result to `ggplot2`. Long format
carries `sample_name`, `band`, and `value`; we add a stable per-spectrum `id`
(sample names repeat across the 5 replicates per group, so grouping lines by
name alone would connect unrelated samples) and the folded group so colors and
legend match the base-R plots above.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 4}
long         = to_long(s)
long$id      = rep(seq_len(nrow(s)), times = ncol(s))
long$species = rep(grp_of, times = ncol(s))

ggplot2::ggplot(long, ggplot2::aes(band, value, group = id, colour = species)) +
    ggplot2::geom_line(alpha = 0.6, linewidth = 0.3) +
    ggplot2::scale_colour_manual(values = group_palette(s)$pal) +
    ggplot2::labs(x = "wavelength (nm)", y = "reflectance", colour = NULL) +
    theme_spectra()
```

Because the long data is a plain data.frame, the rest of the grammar composes
normally. `facet_wrap()` breaks the overplotted bundle into small multiples --
one panel per group, which reads far better than a single crowded axis.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 5}
ggplot2::ggplot(long, ggplot2::aes(band, value, group = id, colour = species)) +
    ggplot2::geom_line(alpha = 0.7, linewidth = 0.3, show.legend = FALSE) +
    ggplot2::scale_colour_manual(values = group_palette(s)$pal) +
    ggplot2::facet_wrap(~ sample_name) +
    ggplot2::labs(x = "wavelength (nm)", y = "reflectance") +
    theme_spectra()
```

The `ggplot2` equivalent of the shaded quantile band is `geom_ribbon()`. Compute
per-band quantiles with `quantile()` -- which returns one spectrum per
probability -- then plot the outer probabilities as a ribbon and the median as a
line.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 4}
qt  = quantile(s, probs = c(0.05, 0.5, 0.95))
qv  = value(qt)
fan = data.frame(band = bands(qt), lo = qv[1, ], mid = qv[2, ], hi = qv[3, ])

ggplot2::ggplot(fan, ggplot2::aes(band)) +
    ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi),
                         fill = spec_pal[1], alpha = 0.25) +
    ggplot2::geom_line(ggplot2::aes(y = mid), colour = spec_pal[1], linewidth = 0.6) +
    ggplot2::labs(x = "wavelength (nm)", y = "reflectance",
                  title = "Median and 5-95% band") +
    theme_spectra()
```

For a one-liner default, `autoplot(s)` draws all spectra as a bare `ggplot`
object you can extend. Note that `autoplot()` drops metadata, so it cannot color
by a grouping on its own -- when you need groups, go through `to_long()` as
above and add the color aesthetic yourself.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 3.5}
ggplot2::autoplot(s) + theme_spectra()
```

## Spectral redundancy: band autocorrelation

Adjacent bands measure almost the same thing, which is why a 2101-band spectrum
carries far fewer than 2101 independent numbers. The band-by-band correlation
matrix makes this concrete. A full-resolution `cor()` would be a 2101x2101
matrix -- needless compute and a multi-megabyte figure -- so we first downsample
to ~10 nm resolution with `resample()`.

```{r, fig.width = 6, fig.height = 5.5}
sc = resample(s, seq(400, 2400, by = 10), fwhm = 10)   # ~200 bands
M  = cor(as.matrix(sc))
wl = bands(sc)

op = graphics::par(mar = c(4, 4, 1, 1))
image(x = wl, y = wl, z = M, zlim = c(-1, 1),
      col = spec_div(255), useRaster = TRUE,
      xlab = "wavelength (nm)", ylab = "wavelength (nm)")
graphics::par(op)
```

The diverging palette is centered at zero (`zlim = c(-1, 1)`): blue is positive
correlation, red negative, pale neutral near zero. The image is dominated by
large blocks of near-perfect positive correlation -- the visible (400-700 nm),
the NIR plateau (~800-1300 nm), and the SWIR regions -- separated by the water
absorption features and the red edge, where correlation drops or flips sign.
This block structure is exactly why a few components can reconstruct a spectrum
almost perfectly, which the next section exploits.

## Dominant axes: PCA

Principal component analysis rotates the band axes into a new set of orthogonal
axes ordered by the variance they explain. Because spectra are so redundant, the
first two or three components typically capture the overwhelming majority of the
variation. We run `prcomp()` on the sample-by-band matrix.

```{r}
pca = prcomp(as.matrix(s), center = TRUE)
imp = summary(pca)$importance
round(imp[, 1:5], 4)
```

```{r}
ve  = round(imp[2, 1:2] * 100, 1)   # % variance explained by PC1, PC2
ve
```

A score scatter of PC1 versus PC2 places each sample in this reduced space;
coloring by group shows how the axes relate to the categorical structure.

```{r, fig.width = 6, fig.height = 5}
lab = paste0(c("PC1", "PC2"), " (", ve, "%)")

op = graphics::par(mar = c(4, 4, 1, 1))
plot(pca$x[, 1], pca$x[, 2], col = species_cols(s), pch = 19,
     xlab = lab[1], ylab = lab[2])
legend("topright", legend = names(gp$pal), col = gp$pal, pch = 19,
       bty = "n", cex = 0.8, ncol = 2)
graphics::par(op)
```

The scores tell you *where* samples sit; the loadings tell you *which
wavelengths* build each axis. Plotting `pca$rotation` over wavelength -- a
"loadings-as-spectra" view -- reveals the spectral fingerprint of each
component. We set up an empty plotting frame, shade the regions, then overlay
the first two loading vectors.

```{r, fig.width = 7, fig.height = 4.5}
rot = pca$rotation[, 1:2]
wl2 = bands(s)

op = graphics::par(mar = c(4, 4, 1, 1))
matplot(wl2, rot, type = "n", xlab = "wavelength (nm)", ylab = "loading")
plot_regions(s, col = spec_region_tints)
graphics::abline(h = 0, col = spec_muted, lty = 3)
graphics::matlines(wl2, rot, col = c(spec_pal[1], spec_pal[6]), lty = 1, lwd = 2)
legend("topright", legend = lab, col = c(spec_pal[1], spec_pal[6]),
       lwd = 2, bty = "n", cex = 0.8)
graphics::par(op)
```

Peaks and troughs in a loading mark the bands that drive that component; a
loading flat near zero means those wavelengths contribute nothing to that axis.
Reading the scores and loadings together -- which samples separate along an
axis, and which wavelengths define it -- is the core of interpreting spectral
variation.

## Data note

`spec_matrix_example` is **simulated**. Real reflectance would show cleaner,
more textbook structure -- sharper water-absorption features, a crisper red
edge, and correlation blocks that align even more tightly with the pigment,
water, and dry-matter absorptions that organize plant spectra. The workflow,
however, is identical on measured data.
